---
title: "Integrating morphology and gene expression with kstitch"
subtitle: "A Xenium Prime melanoma keratinocyte case study"
output:
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 2
vignette: >
  %\VignetteIndexEntry{Integrating morphology and gene expression with kstitch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
<style>
body {
  max-width: 1000px;
}

img {
  border: none !important;
  box-shadow: none !important;
}

body {
  max-width: 1000px;
}
details {
  border: 1px solid #0073e6;        /* Solid blue border */
  background-color: #f8fbff;        /* Very subtle light blue background */
  padding: 12px 15px;               /* Inner spacing so text doesn't touch the box */
  border-radius: 6px;               /* Soft rounded corners */
  margin-bottom: 15px;              /* Space below the box */
}

summary {
  font-weight: bold;
  color: #005bb5;                   /* Darker blue text for the clickable header */
  cursor: pointer;                  /* Changes mouse to a hand icon */
  margin: -12px -15px;              /* Expands the clickable area to the edges */
  padding: 12px 15px;
}

details[open] summary {
  border-bottom: 1px solid #cce5ff; /* Adds a divider line when opened */
  margin-bottom: 10px;
}
</style>

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 5,
  fig.height = 3,
  dev = 'svg',
  fig.retina = 2,
  fig.align = "center",
  out.width = "50%"
)
```

## Core workflow

This section condenses the full vignette into the minimal end-to-end analysis:
load a Xenium object, compute expression factors with NMF, compute TPCA shape
embeddings from nucleus or cell contours, optionally regress technical
covariates, and link morphology to expression with CCA. The detailed sections
below explain and visualize each step.


### Download the Xenium dataset

The vignette uses the 10x Genomics Xenium Prime Human Skin FFPE dataset. Download it using one of the methods below.

### Option 1: R (using `download.file`)

```{r, eval = FALSE}
url <- "https://cf.10xgenomics.com/samples/xenium/3.0.0/Xenium_Prime_Human_Skin_FFPE/Xenium_Prime_Human_Skin_FFPE_outs.zip"
destfile <- "/path/to/Xenium_Prime_Human_Skin_FFPE_outs.zip"
download.file(url, destfile, mode = "wb")
unzip(destfile)
```

### Option 2: Shell (using `curl` or `wget`)

```bash
# Using curl
curl -o Xenium_Prime_Human_Skin_FFPE_outs.zip \
  https://cf.10xgenomics.com/samples/xenium/3.0.0/Xenium_Prime_Human_Skin_FFPE/Xenium_Prime_Human_Skin_FFPE_outs.zip
unzip Xenium_Prime_Human_Skin_FFPE_outs.zip

# Or using wget
wget https://cf.10xgenomics.com/samples/xenium/3.0.0/Xenium_Prime_Human_Skin_FFPE/Xenium_Prime_Human_Skin_FFPE_outs.zip
unzip Xenium_Prime_Human_Skin_FFPE_outs.zip
```

Set `xenium_data_path` to the folder where the zip contents were extracted.
```{r download-data, eval = FALSE}
# Download the Xenium Prime Human Skin FFPE dataset from 10x Genomics and
# unpack it into a local directory. Update xenium_data_path to match.
xenium_data_path <- "/path/to/Xenium_Prime_Human_Skin_FFPE_outs/"
```


```{r core-workflow, eval=FALSE, warning=FALSE}
library(kstitch)
library(Seurat)
library(dplyr)
library(lme4)
library(arrow)

# 1. Load Xenium expression data and segmentation information.
xenium_data_path <- "/path/to/Xenium_Prime_Human_Skin_FFPE_outs/"

xenium_obj <- LoadXenium(
  xenium_data_path,
  molecule.coordinates = FALSE
)

segmentation_info <- ReadXenium(
  xenium_data_path,
  outs = "segmentation_method",
  type = c("segmentations", "nucleus_segmentations", "centroids")
)

# Optional: subset to the cells and genes used in this vignette.
analysis_obj <- subset(
  xenium_obj,
  cells = keratinocyte_cell_ids,
  features = keratinocyte_genes_to_use
)

# 2. Compute NMF expression factors and store them in the Seurat object.
nmf_output <- compute_nmf(
  analysis_obj,
  assay_name               = "Xenium",
  stability_seed           = 235,
  default_num_factors      = 10,
  num_top_genes_per_factor = 50,
  integration_method       = "online",
  selectGenes_params       = list(nGenes = "all", verbose = TRUE),
  gene_pct_threshold       = 0.01
)

analysis_obj <- store_nmf_results(
  analysis_obj,
  nmf_output,
  reduction_prefix = "nmf"
)

# 3. Choose one contour type and compute TPCA morphology embeddings.
# For nuclei, keep cells with exactly one nucleus. For whole-cell contours,
# use Cells(analysis_obj) directly.
cell_metadata <- arrow::read_parquet(
  file.path(xenium_data_path, "cells.parquet")
)

single_nucleus_cells <- cell_metadata |>
  filter(nucleus_count == 1) |>
  pull(cell_id) |>
  intersect(Cells(analysis_obj))

contour_type <- "nucleus"  # change to "cell" for whole-cell morphology
contour_cells <- if (contour_type == "nucleus") {
  single_nucleus_cells
} else {
  Cells(analysis_obj)
}

tpca_result <- run_tpca_from_seurat(
  xenium_obj,
  seg_list      = segmentation_info,
  output_dir    = file.path("output", paste0("tpca_", contour_type)),
  contour_type  = contour_type,
  use_parallel  = TRUE,
  num_threads   = 8,
  cell_ids      = contour_cells,
  store_history = TRUE
)

shape_reduction <- paste0("tpca_", contour_type)
shape_key <- if (contour_type == "nucleus") {
  "NucleusShapePC_"
} else {
  "CellShapePC_"
}

analysis_obj <- store_tpca_results(
  analysis_obj,
  tpca_result,
  reduction_key_prefix = shape_key
)

# 4. Build matched morphology and expression matrices.
shape_embeddings <- Embeddings(analysis_obj, shape_reduction)
expr_embeddings  <- Embeddings(analysis_obj, "nmf")
shared_cells     <- intersect(rownames(shape_embeddings), rownames(expr_embeddings))

shape_mat <- shape_embeddings[shared_cells, 1:10, drop = FALSE]
expr_mat  <- expr_embeddings[shared_cells, , drop = FALSE]

log_area <- setNames(
  log(tpca_result$Metadata$area[
    match(shared_cells, tpca_result$Metadata$cell_id)
  ]),
  shared_cells
)

shape_mat <- cbind(
  shape_mat,
  scaled_log_area = as.numeric(scale(log_area[shared_cells]))
)

# 5. Optional: regress sequencing depth and segmentation method.
.regress_lmer <- function(mat, scaled_log_depth, seg_method) {
  cell_ids <- rownames(mat)
  out <- matrix(
    NA_real_,
    nrow = nrow(mat),
    ncol = ncol(mat),
    dimnames = dimnames(mat)
  )

  for (j in seq_len(ncol(mat))) {
    df <- data.frame(
      y = mat[, j],
      scaled_log_depth = scaled_log_depth[cell_ids],
      seg_method = seg_method[cell_ids]
    )
    ok <- complete.cases(df)
    if (sum(ok) < 10L) next

    fit <- lmer(
      y ~ scaled_log_depth + (1 | seg_method),
      data = df[ok, ],
      REML = FALSE
    )
    out[ok, j] <- residuals(fit)
  }

  out
}

meta <- analysis_obj@meta.data
log_depth <- setNames(log(meta$nCount_Xenium), rownames(meta))
seg_method <- setNames(meta$segmentation_method, rownames(meta))
scaled_log_depth <- setNames(
  scale(log_depth[shared_cells])[, 1],
  shared_cells
)

shape_mat_reg <- .regress_lmer(shape_mat, scaled_log_depth, seg_method)
expr_mat_reg  <- .regress_lmer(expr_mat, scaled_log_depth, seg_method)

keep <- complete.cases(shape_mat_reg) & complete.cases(expr_mat_reg)
shape_mat_reg <- shape_mat_reg[keep, , drop = FALSE]
expr_mat_reg  <- expr_mat_reg[keep, , drop = FALSE]

# 6. Run CCA and store the linked morphology-expression representation.
cca_result <- link_shape_and_factors(
  obj       = analysis_obj,
  shape_mat = shape_mat_reg,
  expr_mat  = expr_mat_reg
)

reduction_name <- paste0(contour_type, "_cca_regressed")
analysis_obj <- store_kstitch_results(
  analysis_obj,
  cca_result,
  reduction_name    = reduction_name,
  reduction_key_csp = "CSPReg_",
  reduction_key_cep = "CEPReg_"
)

# 7. Inspect the main outputs.
cca_result$CC_Corr_Coefs
plot_shape_modes(tpca_result, sds_to_plot = c(-2, -1, 0, 1, 2), num_pcs = 4L)
plot_csp_loadings(cca_result, cc_idx = 1)
plot_cep_loadings(cca_result, cc_idx = 1)
plot_csp_boundary_montage(
  group_result           = cca_result,
  tpca_result            = tpca_result,
  score_type             = "CSP",
  cc_idx                 = 1L,
  num_bins               = 10L,
  num_cells_for_boundary = 5L,
  norm_rescale           = TRUE
)
```

---

## Introduction

Spatial transcriptomics tells us *where* genes are expressed, but the shape of
a cell — how round it is, how large its nucleus, how irregular its boundary —
carries independent biological information that transcript counts alone cannot
capture. **kstitch** links these two modalities by computing a low-dimensional
representation of cell morphology (via Tangent PCA, or TPCA) and finding the
axes of maximal correlation between shape and gene expression (via canonical
correlation analysis, or CCA).

This vignette walks through a complete kstitch analysis on a subset of
keratinocytes from a human melanoma Xenium Prime dataset. The object used here
has been pre-computed (TPCA, NMF, and CCA have already been run and stored
inside the object) so that the vignette can be reproduced without Python or
`rliger`. The script that built this object lives in
`data-raw/prepare_vignette_data.R`.

### What is TPCA?

Standard PCA operates on flat Euclidean space. Cell contour polygons live on a
curved manifold (Kendall's shape space), where straight-line interpolation
between shapes is geometrically meaningless. Tangent PCA (also called Principal
Geodesic Analysis, or PGA) works by:

1. Finding the **Fréchet mean** — the average shape in the curved space.
2. Projecting all shapes onto the **tangent plane** at that mean via a log-map.
3. Running standard SVD on the projected (now flat) coordinates.

The result is a set of **shape principal components** (Shape PCs) that describe
the major axes of morphological variation in a geometrically correct way.

### What does CCA find?

CCA finds linear combinations of the shape PCs (called **CSP** — Canonical
Shape variates) and linear combinations of the NMF factors (called **CEP** —
Canonical Expression variates) such that each CSP/CEP pair has maximal
correlation. A high canonical correlation on the first component means there is
a strong axis of joint morpho-molecular variation: cells that differ in shape
along CSP1 also differ in gene expression along CEP1.

---

## Setup

```{r load-libs, warning=FALSE}
library(kstitch)
library(Seurat)
library(ggplot2)
library(dplyr)

ggplot2::theme_set(
  ggplot2::theme_minimal() +
    ggplot2::theme(
      panel.background = ggplot2::element_blank(),
      panel.border = ggplot2::element_blank(),
      plot.background = ggplot2::element_blank()
    )
)
```

Load the pre-computed vignette object. If you ran the full pipeline chunks
below, `keratinocyte_obj` will already be in memory and you can skip this chunk.

```{r load-obj, warning=FALSE}
vignette_data         <- readRDS(system.file("extdata", "kstitch_vignette_xenium.rds",
                                             package = "kstitch"))
vignette_obj      <- vignette_data$obj
tpca_nucleus_metadata <- vignette_data$metadata$tpca_nucleus
tpca_cell_metadata    <- vignette_data$metadata$tpca_cell

tpca_cell_info <- readRDS(system.file("extdata", "kstitch_vignette_tpca_cell.rds",
                                             package = "kstitch") )

tpca_nucleus_info <- readRDS(system.file("extdata", "kstitch_vignette_tpca_nucleus.rds",
                                             package = "kstitch") )

keratinocyte_obj <- vignette_obj

melanoma_xenium_segmentation_info <- readRDS(system.file("extdata", "kstitch_keratinocyte_segmentations.rds",
                                             package = "kstitch"))
```

The object contains 3,171 keratinocytes. 2,331 have nucleus TPCA shape PCs
computed and 3,171 have cell TPCA shape PCs computed. Cells missing from the nuclearTPCA
reduction had missing nuclear contours and were excluded from
the TPCA run. Below, we list all the object reductions available via Seurat.

```{r check-reductions}
names(vignette_obj@reductions)
sum(!is.na(Seurat::Embeddings(vignette_obj, "tpca_nucleus")[, 1]))
sum(!is.na(Seurat::Embeddings(vignette_obj, "tpca_cell")[, 1]))
```

---

## Step 1 — Download data and run NMF (eval=FALSE)

NMF is run once and its factors feed into both the nuclear and cell shape
analyses below.

Note that the link above points to the full Xenium output bundle and not the Xenium Explorer subset. Once downloaded and unzipped, the vignette assumes the extracted directory is available locally. The file tree in the extracted directory should contain the following files --

<pre style="font-family: monospace;">
Xenium_Prime_Human_Skin_FFPE_outs/
├── analysis_summary.html
├── analysis.tar.gz
├── analysis.zarr.zip
├── aux_outputs.tar.gz
├── cell_boundaries.csv.gz
├── <b>cell_boundaries.parquet</b>
├── cell_feature_matrix.h5
├── cell_feature_matrix.tar.gz
├── cell_feature_matrix.zarr.zip
├── cell_feature_matrix/
│   ├── barcodes.tsv.gz
│   ├── features.tsv.gz
│   └── matrix.mtx.gz
├── cells.csv.gz
├── <b>cells.parquet</b>
├── cells.zarr.zip
├── experiment.xenium
├── gene_panel.json
├── metrics_summary.csv
├── morphology.ome.tif
├── morphology_focus/
│   ├── morphology_focus_0000.ome.tif
│   ├── morphology_focus_0001.ome.tif
│   ├── morphology_focus_0002.ome.tif
│   └── morphology_focus_0003.ome.tif
├── nucleus_boundaries.csv.gz
├── <b>nucleus_boundaries.parquet</b>
├── transcripts.parquet
└── transcripts.zarr.zip
</pre>

They key files the pipeline relies on (marked in bold) are `nuclear_boundaries.parquet` and `cell_boundaries.parquet`, which contain `(x,y)` coordinates of nuclear and cellular contours, respectively, from the on-board Xenium analyzer segmentation pipeline. The `csv.gz` files contain identical information to the boundaries parquet files but we rely on the latter for speed. `cells.parquet` contains two key pieces of information --- the number of segmented nuclei in a cell and the cellular segmentation method employed. The latter is relevant for later Xenium protocols that rely on multi-modal staining to infer cellular boundaries. 

### Load the Xenium object and subset to keratinocytes

```{r subset-keratinocytes, eval = FALSE, warning = FALSE}
keratinocyte_cell_ids     <- Cells(vignette_obj)
keratinocyte_genes_to_use <- Features(vignette_obj)  # genes >= 10 nTPM in keratinocytes profiled in human protein atlas (HPA). Vignette object subsetted. 

melanoma_xenium_obj <- LoadXenium(xenium_data_path, molecule.coordinates = FALSE) #Creates Seurat object but without nuclear and cellular segmentation contours loaded. Resulting Seurat object has read counts loaded in the "Xenium" assay by default. 
keratinocyte_obj <- subset(melanoma_xenium_obj,
                           cells    = keratinocyte_cell_ids,
                           features = keratinocyte_genes_to_use)
# You may receive object-validation warnings; these can be ignored.
```

### Run NMF

`compute_nmf` runs integrative NMF from the `rliger` package. The resulting output can be stored in a Seurat object using `store_nmf_results` which stores factor embeddings as a `DimReduc` object. This allows clustering and other visualizations to be computed using existing functions in `Seurat`. 


```{r run-nmf, warning = FALSE}
num_factors <- 10

nmf_output <- compute_nmf(
  keratinocyte_obj,
  assay_name               = "Xenium",
  stability_seed           = 235,
  default_num_factors      = num_factors,
  num_top_genes_per_factor = 50,
  integration_method       = "online",
  selectGenes_params       = list(nGenes = "all", verbose = TRUE),
  gene_pct_threshold       = 0.01
)

#Stores NMF results in the "nmf" reductions slot
keratinocyte_obj <- store_nmf_results(keratinocyte_obj, nmf_output, reduction_prefix = "nmf")
```

<details>
<summary><strong>Optional workflow to analyze stored NMF output in Seurat object</strong></summary>
```{r optional-nmf-seurat-workflow, fig.height = 3, out.width="50%"}
#Gene loadings on each factor can be accessed using Loadings(), which is also accessible via nmf_output$NMF_Loading. 
nmf_loading_mat <- Loadings(keratinocyte_obj, reduction="nmf")

#Top K genes per factor can be extracted using the TopFeatures function in SeuratObject. The snippet below
#extracts top K = 50 genes with the highest loading in each factor. 
nmf_loadings <- SeuratObject::Loadings(keratinocyte_obj[["nmf"]])

nmf_top_features <- lapply(
  seq_len(ncol(nmf_loadings)),
  function(i) {
    Seurat::TopFeatures(
      keratinocyte_obj[["nmf"]],
      dim = i,
      nfeatures = 50
    )
  }
)

names(nmf_top_features) <- colnames(nmf_loadings)

keratinocyte_obj <- FindNeighbors( keratinocyte_obj, reduction="nmf",verbose=F) %>% FindClusters(.,verbose=F) %>% RunUMAP(.,dims=1:num_factors,reduction = "nmf",verbose=F)

VlnPlot( keratinocyte_obj, features="NMF_1")

FeaturePlot( keratinocyte_obj,  features="NMF_1", reduction="umap")
```
</details>

### Load cellular and nuclear contours using ReadXenium

The `ReadXenium` command from the `Seurat` package loads nuclear contours (`nuclear_segmentations`), cellular contours (`segmentations`) and the location of each cell (`centroids`) from the downloaded Xenium data bundle. `segmentation_method` contains information on how each given cell's contour was determined. Pass only type = c("nucleus_segmentations") to load only nuclear contours and type = c("segmentations") to load only cellular contours. 

```{r load-seg, eval = FALSE}
melanoma_xenium_segmentation_info <- ReadXenium(
  xenium_data_path,
  outs = c("segmentation_method"),
  type = c("segmentations", "nucleus_segmentations", "centroids") 
)
```

---

# Nuclear shape analysis

## Step 2 — Run TPCA on nuclear contours. 

TPCA is restricted to cells that have exactly one nucleus. Including cells with multiple nuclear contours can skew the shape modes computed by the method and make the intepretation of shape modes challenging. Information on which cells have a single nucleus assigned to them can be gleaned from the `nucleus_count` column in the cells.parquet` parquet file . Reading parquet files requires the `arrow` library in R. If installing `arrow` is an issue, the `cells.csv.gz` can be read using `read.csv`. 
```{r single-nucleus-filter, eval = FALSE}
library(arrow)
library(dplyr)
cell_seg_meta_data_df <- read_parquet( file.path( xenium_data_path, "cells.parquet"))

cell_ids_with_single_nucleus <- dplyr::filter( cell_seg_meta_data_df, nucleus_count == 1) %>% pull(cell_id)
```

Below, we use keratinocyte cell IDs from data shipped with the vignette. 

```{r run-tpca-nucleus, eval = FALSE}
keratinocytes_cells_with_single_nucleus <-
  rownames(na.omit(Embeddings(vignette_obj,"tpca_nucleus")))

tpca_nucleus_info <- run_tpca_from_seurat(
  melanoma_xenium_obj,
  seg_list     = melanoma_xenium_segmentation_info,
  output_dir   = "output/tpca_nucleus",    #Optional. If omitted, data is written to a temporary directory. 
  contour_type = "nucleus", 
  use_parallel = TRUE,
  num_threads  = 8,
  cell_ids     = keratinocytes_cells_with_single_nucleus, #Optional. For sub-setting nuclei. 
  store_history = TRUE    #Optional. Allows change in estimate of mean shape to be tracked along with convergence towards mean. 
)

#You may receive a warning about some cells having NA scores in their TPCA embedding. This reflects 
#the fact that cells with zero or more than one nuclear contour were omitted from TPCA embedding in the previous step. 
#TPCA embeddings are stored under the reduction name "tpca_nucleus" by default. 
keratinocyte_obj <- store_tpca_results(keratinocyte_obj, tpca_nucleus_info,
                                       reduction_key_prefix = "NucleusShapePC_")
```
### Analyzing TPCA output

**Visualizing shape variation** : `plot_shape_modes()` reconstructs nucleus shapes along each principal geodesic
component at a range of standard deviation values to
understand what mode of variation each shape PC captures. 
In the plow below, each row is a shape PC and each column is a standard deviation along that component. 

```{r shape-modes-nuc, fig.height = 3, out.width = "50%"}
plot_shape_modes(
  tpca_nucleus_info,
  sds_to_plot = c(-2, -1, 0, 1, 2),
  num_pcs     = 4L
)
```

**Convergence to mean** - In the event that the mean shape (shown at $\sigma$ = 0 in the plot above) does not resemble a nucleus at all, `plot_mu_history` allows one to visualize successive estimates of the mean during convergence to the final estimate. The raw matrices representing coordinates of the Frechet mean shape can be retrieved from `attr(tpca_nucleus_info,'frechet_history')`and can be plotted using the `plot_mu_history` function. In the example below, the estimate of the mean shape is shown at every $10^{th}$ iteration starting from the first one. 

```{r frechet-mean-estimates, fig.height = 3, out.width = "50%"}
plot_mu_history(
  tpca_nucleus_info,n = 10,linewidth=0.5
)
```

The change in the estimate of the mean shape with each iteration can be plotted using the `plot_frechet_convergence` function. When no tolerance is specified using the `frechet_mean_tol` parameter, iterations are terminated when the norm of the tangent vector is $< 10^{-4}$.  

```{r frechet-mean-convergence, fig.height = 3, out.width = "60%"}
plot_frechet_convergence(
  tpca_nucleus_info
)
```

<details>
<summary><strong>Optional workflow to cluster and analyze stored TPCA output in Seurat object</strong></summary>
```{r optional-tpca-seurat-workflow, out.width="50%", fig.height=3}
#TPCA embedding on each factor can be accessed using Loadings(), which is also accessible via nmf_output$NMF_Loading. Since
#some cells have NA in their TPCA embedding, we keep track of cells that have valid TPCA embeddings associated with them.
#Note that this step is not needed if there are no NA values present. 
tpca <- SeuratObject::Embeddings(keratinocyte_obj,reduction="tpca_nucleus")
top_shape_pcs_to_use <- 10

valid_cells <- rownames(tpca)[
  apply(tpca[, 1:top_shape_pcs_to_use, drop = FALSE], 1, function(x) all(is.finite(x)))
]

#Alternately, TPCA embeddings can be accessed directly from output of run_tpca_from_seurat. Note that column names
#will differ from those of embeddings retrieved using Embeddings on the Seurat object.
tpca_embedding_mat <- tpca_nucleus_info$TPCA_Embedding

#Remove NA cells and compute UMAP embeddings and clusters based on TPCA embeddings.
top_shape_pcs_to_use <- 10

keratinocyte_obj <- subset(keratinocyte_obj, cells = valid_cells) |>
  Seurat::FindNeighbors(reduction = "tpca_nucleus", dims = 1:top_shape_pcs_to_use) |>
  Seurat::FindClusters() |>
  Seurat::RunUMAP(reduction = "tpca_nucleus", dims = 1:top_shape_pcs_to_use)

VlnPlot( keratinocyte_obj, features="NucleusShapePC_3")

FeaturePlot( keratinocyte_obj,  features="NucleusShapePC_1", reduction="umap")
```
</details>

## Concatenating nuclear area to shape
Though TPCA explicitly quotients out nuclear size prior to registering nuclear contour shapes and performing PCA, size represents a biologically meaningful variable. Below, we concatenate the log of nuclear area to the top ten nuclear shape PCs of each cell to create a joint representation of each nucleus based on its size (i.e., area) and shape. 

```{r nucleus-area-concatenation}
tpca <- SeuratObject::Embeddings(keratinocyte_obj,reduction="tpca_nucleus")
top_shape_pcs_to_use <- 10

valid_cells <- rownames(tpca)[
  apply(tpca[, 1:top_shape_pcs_to_use, drop = FALSE], 1, function(x) all(is.finite(x)))
]

nuc_log_area_vec <- setNames(
  log(tpca_nucleus_info$Metadata$area[
        match(valid_cells, tpca_nucleus_info$Metadata$cell_id)]),
 valid_cells
)
top_shape_pcs_to_use <- 10

tpca_mat  <- Embeddings(keratinocyte_obj, "tpca_nucleus")[valid_cells, 1:top_shape_pcs_to_use]
tpca_with_nuc_log_area_mat <- cbind(tpca_mat,
                              scaled_log_area = as.numeric(scale(nuc_log_area_vec[valid_cells])))

nmf_mat <- Embeddings(keratinocyte_obj, "nmf")
```

## Step 3 — Regression of covariates from NMF and TPCA (and area). 
The final step of the STITCH pipeline involves computing a CCA between the joint nuclear shape PC and area matrix
on the one hand with the factor embedding matrix on the other. We describe an optional regression step below
prior to CCA which may be warranted depending on the study or to take into account technical covariates. 

For instance, variation in total RNA counts per cell across an image may reflect both technical and biological variation. In the case of a Xenium dataset, one such source of technical variation is the choice of segmentation method used to determine cell boundaries (in the case of Xenium multimodal segmentation). The segmentation method used to draw a given cell's contours influences the area assigned to the cell and the number of transcripts assigned to it. Below, we define a helper function for 
regressing total RNA count and segmentation method from shape PCs, nuclear area and each factor's score. 


### Covariate regression helper

We define `.regress_lmer()` once here for use in both the nuclear and cell
contour analyses. It operates column-by-column on a matrix: for each column it
fits a mixed-effects model with `scaled_log_depth` (z-score of log of RNA counts per cell) 
as a fixed effect and `segmentation_method` as a random intercept, then returns the model residuals.
Cells with incomplete covariate data are excluded from model fitting and left
as `NA` in the output.

```{r define-regress-lmer}
library(lme4)

.regress_lmer <- function(mat, scaled_log_depth, seg_method) {
  cell_ids  <- rownames(mat)
  resid_mat <- matrix(NA_real_, nrow = nrow(mat), ncol = ncol(mat),
                      dimnames = dimnames(mat))
  for (j in seq_len(ncol(mat))) {
    df <- data.frame(
      y                = mat[, j],
      scaled_log_depth = scaled_log_depth[cell_ids],
      seg_method       = seg_method[cell_ids],
      stringsAsFactors = FALSE
    )
    ok <- complete.cases(df)
    if (sum(ok) < 10L) {
      warning(sprintf("Column %d: fewer than 10 complete cases, skipping.", j))
      next
    }
    fit <- lmer(y ~ scaled_log_depth + (1 | seg_method),
                data = df[ok, ], REML = FALSE)
    resid_mat[ok, j] <- residuals(fit)
  }
  resid_mat
}
```

We use this function to carry out the regression of both these covariates from `tpca_with_nuc_log_area_mat` and
`nmf_mat`. 

```{r nucleus-regression}
meta           <- keratinocyte_obj@meta.data
log_depth_vec  <- setNames(log(meta$nCount_Xenium),  rownames(meta)) #nCount_Xenium is computed by Seurat in the LoadXenium step.
seg_method_vec <- setNames(meta$segmentation_method, rownames(meta))
nmf_cells      <- rownames(nmf_mat)

tpca_nuc_cells   <- rownames(tpca_with_nuc_log_area_mat)
shared_nuc_cells <- intersect(tpca_nuc_cells, nmf_cells)

scaled_log_depth_nuc <- setNames(scale(log_depth_vec[shared_nuc_cells])[, 1],
                                  shared_nuc_cells)

shape_pcs_nuc       <- tpca_with_nuc_log_area_mat[shared_nuc_cells,]
expr_mat_nuc        <- nmf_mat[shared_nuc_cells, ]

shape_mat_nuc_reg <- .regress_lmer(shape_pcs_nuc,
                                    scaled_log_depth_nuc, seg_method_vec)
expr_mat_nuc_reg  <- .regress_lmer(expr_mat_nuc,
                                    scaled_log_depth_nuc, seg_method_vec)

```


## Step 4 — CCA between nuclear shape PCs and size with factors derived via NMF


```{r nucleus-cca}
keep <- complete.cases(shape_mat_nuc_reg) &
        complete.cases(expr_mat_nuc_reg)

#Eliminate NAs from input matrices
shape_mat_nuc_reg <- shape_mat_nuc_reg[keep, , drop = FALSE]
expr_mat_nuc_reg  <- expr_mat_nuc_reg[keep, , drop = FALSE]

cca_nuc_reg <- link_shape_and_factors(
  obj       = keratinocyte_obj,
  shape_mat = shape_mat_nuc_reg,
  expr_mat  = expr_mat_nuc_reg
)

#CCA without covariate regression 
# cca_nuc_unreg <- link_shape_and_factors(
#   obj       = keratinocyte_obj,
#   shape_mat = shape_pcs_nuc,
#   expr_mat  = expr_mat_nuc
# )

#Replace cca_nuc_reg with cca_nuc_unreg to store CCA results without regression. 
keratinocyte_obj <- store_kstitch_results(keratinocyte_obj, cca_nuc_reg,
                                           reduction_name    = "nuc_cca_regressed",
                                           reduction_key_csp = "CSPAfterReg_",
                                           reduction_key_cep = "CEPAfterReg_")
```

<details open>
<summary><strong>Optional workflow to multi-modally cluster and analyze stored CCA output in Seurat object</strong></summary>
Both CSP and CEP matrices can be obtained from `keratinocyte_obj` via the `Embedding` command or directly from the output of `link_shape_and_factors` via `cca_nuc_reg$CSP_Scores` and `cca_nuc_reg$CEP_Scores`. Below, we demonstrate a basic multi-modal clustering scheme with Seurat based on the top 3 CEPs and CSPs. 

```{r optional-cca-seurat-workflow, out.width="50%", fig.height=3}
nuc_csp_mat <- SeuratObject::Embeddings(keratinocyte_obj,reduction="nuc_cca_regressed_csp")
nuc_cep_mat <- SeuratObject::Embeddings(keratinocyte_obj,reduction="nuc_cca_regressed_cep")

top_csps_to_use <- 3
top_ceps_to_use <- 3
nuc_csp_cep_mat <- cbind( nuc_csp_mat[,1:top_csps_to_use], nuc_cep_mat[1:top_ceps_to_use])
colnames(nuc_csp_cep_mat) <- paste("CC",1:ncol(nuc_csp_cep_mat))

dims_to_use <- 1:ncol(nuc_csp_cep_mat)
#The first three columns of the dimensionality reduction are CSPs 1-3 while the last three are CEPs 1-3
external_cep_csp_dim_reduc <- CreateDimReducObject(
    embeddings = nuc_csp_cep_mat,
    key = "CSPCEP",
    assay = "Xenium" 
)
keratinocyte_obj[["combined_csp_cep"]] <- external_cep_csp_dim_reduc

keratinocyte_obj <- subset(keratinocyte_obj, cells = valid_cells) |>
  Seurat::FindNeighbors(reduction = "combined_csp_cep", dims = 1:ncol(nuc_csp_cep_mat),verbose=F) |>
  Seurat::FindClusters(verbose=F,resolution=0.3) |>
  Seurat::RunUMAP(reduction = "combined_csp_cep", dims = 1:ncol(nuc_csp_cep_mat),verbose=F)

DimPlot(keratinocyte_obj)

#Example command to overlay NucleusShapePC_1 and CSPCEP_1 (which is equal to CSPAfterReg_1) on UMAP coordinates
#FeaturePlot( keratinocyte_obj,  features="NucleusShapePC_1", reduction="umap")
#FeaturePlot( keratinocyte_obj,  features="CSPCEP_1", reduction="umap")

```
</details>

### Visualizing canonical correlations

The canonical correlation coefficients associated with CSP and CEP vectors can be accessed via `cca_nuc_reg$CC_Corr_Coefs` (or alternately, `keratinocyte_obj@misc$kstitch$nuc_cca_regressed$CC_Corr_Coefs` when `store_kstitch_results` is run). 
```{r cc-barplot-nuc, fig.height = 4}
k <- length(cca_nuc_reg$CC_Corr_Coefs)

cc_df <- data.frame(
  Component   = paste0("CC", seq_len(k)),
  Correlation = cca_nuc_reg$CC_Corr_Coefs[seq_len(k)]
) %>% mutate(Component=factor(Component,levels=paste0("CC", seq_len(k))))

ggplot(cc_df, aes(x = Component, y = Correlation)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Nucleus CCA — canonical correlations",
       y     = "Canonical correlation") +
  theme_classic()
```

#### CSP loadings

The CSP loading plot shows which shape features load most strongly on each
canonical shape variate. By convention (enforced by `anchor_cca_signs()`),
the sign of each component is set so that the shape feature with the highest
absolute structure correlation loads positively. Below, we see that nuclear area (`scaled_log_area`) contributes more strongly to CSP1 than CSP2 

```{r csp-loadings-nuc-reg-1, out.width = "50%"}
plot_csp_loadings(cca_nuc_reg, cc_idx = 1)
```
```{r csp-loadings-nuc-reg-2, out.width = "50%"}
plot_csp_loadings(cca_nuc_reg, cc_idx = 2)
```

#### CEP loadings

The CEP loading plot shows which NMF factors are most associated with each
morphological axis. Factors with large positive loadings on CEP1 are
up-regulated in cells with high CSP1 scores. To identify the genes with the highest weight in
each NMF factor, inspect the matrix returned by `Loadings(keratinocyte_obj,"nmf")` or
`nmf_output$NMF_Loading`. 

```{r cep-loadings-nuc-reg-1, out.width = "50%"}
plot_cep_loadings(cca_nuc_reg, cc_idx = 1)
```

```{r cep-loadings-nuc-reg-2, out.width = "50%"}
plot_cep_loadings(cca_nuc_reg, cc_idx = 2)
```

#### Variation in nuclear size and shape with CSP/CEP score.

The boundary montage is the headline figure of kstitch. Cells are binned by
their CSP score into deciles. Within each bin the cells closest to the bin
centre (medoids, measured by pairwise TPCA distance) are selected, their
contours are Procrustes-aligned to the Fréchet mean, and displayed. With
`norm_rescale = TRUE`, contours are rescaled by the contour's Frobenius norm
so that the displayed size reflects true nuclear area.

When the pre-shape embeddings (`tpca_nucleus_info$Info$pre_shape_embedding`) is stored,
then `plot_csp_boundary_montage` plots pre-shapes and multiplies them by their Frobenius norm
to plot actual nuclear shapes (as is the case in the vignette). In case pre-shape embeddings are not stored in the output of `run_tpca_from_seurat`, which is a choice left to the user since they can take up a lot of storage in memory depending on the number of contours in the dataset, then `plot_csp_boundary_montage` uses PCs (set by default as the top 10 PCs) to reconstruct nuclear shapes. 

`plot_csp_boundary_montage` groups cells into bins based on their CSP or CEP score (based on `score_type`) along a given CC axis (specified by `cc_idx`). Below, we plot keratinocyte nuclei based on their CSP1 scores. Nuclei in Bin 10 have the highest CSP1 scores and those in Bin 1 have the lowest CSP1 scores. 

```{r montage-nuc-reg-csp, out.width = "75%"}
plot_csp_boundary_montage(
  group_result           = cca_nuc_reg,
  score_type = "CSP",
  tpca_result            = tpca_nucleus_info,
  cc_idx                 = 1L,
  num_bins               = 10L,
  num_cells_for_boundary = 5L,
  norm_rescale           = TRUE
)
```

Alternately, nuclei can be visualized by binning their CEP1 scores into deciles, with nuclei in Bin 10 having the highest CEP1 scores while those in Bin 1 have the lowest CEP1 scores.

```{r montage-nuc-reg-cep, fig.height = 3.5, out.width = "75%"}
plot_csp_boundary_montage(
  group_result           = cca_nuc_reg,
  reconstruction_pcs = 1:10,
  score_type = "CEP",
  tpca_result            = tpca_nucleus_info,
  cc_idx                 = 1L,
  num_bins               = 10L,
  num_cells_for_boundary = 5L,
  norm_rescale           = TRUE
)
```

---

# Cell shape analysis

The cell shape analysis follows exactly the same structure as the nuclear shape
analysis above. The key differences are that contours are drawn from the whole
cell boundary rather than the nucleus, all keratinocytes are included (not just
those with a single nucleus), and the CCA results are stored under
`cell_cca_regressed` in the Seurat object. Since the NMF from Step 1 in the nuclear shape analysis is reused for downstream analyses, we demonstrate step 2 onwards below --


## Step 2 — Export cell contours and run TPCA 

```{r run-tpca-cell}
tpca_cell_info <- run_tpca_from_seurat(
  melanoma_xenium_obj,
  seg_list     = melanoma_xenium_segmentation_info,
  output_dir   = "output/tpca_cell",
  contour_type = "cell",
  use_parallel = TRUE,
  num_threads  = 8,
  cell_ids     = Cells(keratinocyte_obj),
  store_history = TRUE
)
keratinocyte_obj <- store_tpca_results(keratinocyte_obj, tpca_cell_info,
                                       reduction_key_prefix = "CellShapePC_")
```

## Step 3 — Cell covariate regression 

```{r cell-regression}
tpca_cell_cells   <- rownames(Embeddings(keratinocyte_obj, "tpca_cell"))
shared_cell_cells <- intersect(tpca_cell_cells, nmf_cells)

cell_log_area_vec    <- setNames(
  log(tpca_cell_info$Metadata$area[
        match(rownames(meta), tpca_cell_info$Metadata$cell_id)]),
  rownames(meta)
)
scaled_cell_log_area  <- setNames(scale(cell_log_area_vec[shared_cell_cells])[, 1],
                                   shared_cell_cells)
scaled_log_depth_cell <- setNames(scale(log_depth_vec[shared_cell_cells])[, 1],
                                   shared_cell_cells)

shape_pcs_cell       <- Embeddings(keratinocyte_obj, "tpca_cell")[shared_cell_cells, 1:10]
expr_mat_cell        <- Embeddings(keratinocyte_obj, "nmf")[shared_cell_cells, ]
shape_mat_cell_unreg <- cbind(shape_pcs_cell,
                               scaled_log_area = scaled_cell_log_area[shared_cell_cells])

shape_mat_cell_reg <- .regress_lmer(shape_mat_cell_unreg,
                                     scaled_log_depth_cell, seg_method_vec)
expr_mat_cell_reg  <- .regress_lmer(expr_mat_cell,
                                     scaled_log_depth_cell, seg_method_vec)
```

## Step 4 — Cell CCA

```{r cell-cca}
#CCA without covariate regression
cca_cell_unreg <- link_shape_and_factors(
  obj       = keratinocyte_obj,
  shape_mat = shape_mat_cell_unreg,
  expr_mat  = expr_mat_cell
)

cca_cell_reg <- link_shape_and_factors(
  obj       = keratinocyte_obj,
  shape_mat = shape_mat_cell_reg,
  expr_mat  = expr_mat_cell_reg
)

#Replace cca_cell_reg with cca_cell_unreg to store CCA results without regression. 
keratinocyte_obj <- store_kstitch_results(keratinocyte_obj, cca_cell_reg,
                                           reduction_name    = "cell_cca_regressed",
                                           reduction_key_csp = "CSPCR_",
                                           reduction_key_cep = "CEPCR_")

#Command below helps recover CCA results from Seurat object
cell_cca_reg   <- get_kstitch_results(keratinocyte_obj, "cell_cca_regressed")

```

## Cell shape results

Retrieve the cell CCA results and build the `tpca_result` object. If you ran
the full pipeline, use `tpca_cell_info$Metadata` in place of
`tpca_cell_metadata`.

```{r retrieve-cell-cca}
```

### Shape modes

```{r shape-modes-cell, fig.height = 6}
plot_shape_modes(
  tpca_cell_info,
  sds_to_plot = c(-2, -1, 0, 1, 2),
  num_pcs     = 4L
)
```

### Canonical correlations

```{r cc-barplot-cell, fig.height = 4}
k <- length(cell_cca_reg$CC_Corr_Coefs)

cc_df <- data.frame(
  Component   = paste0("CC", seq_len(k)),
  Correlation = cell_cca_reg$CC_Corr_Coefs[seq_len(k)]
) %>% mutate(Component=factor(Component,levels=paste0("CC", seq_len(k))))


ggplot(cc_df, aes(x = Component, y = Correlation)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Cell CCA — canonical correlations",
       y     = "Canonical correlation") +
  theme_classic()
```

### CSP loadings

```{r csp-loadings-cell-reg-1}
plot_csp_loadings(cell_cca_reg, cc_idx = 1)
```

```{r csp-loadings-cell-reg-2}
plot_csp_loadings(cell_cca_reg, cc_idx = 2)
```

### CEP loadings

```{r cep-loadings-cell-reg-1}
plot_cep_loadings(cell_cca_reg, cc_idx = 1)
```

```{r cep-loadings-cell-reg-2}
plot_cep_loadings(cell_cca_reg, cc_idx = 2)
```

### CSP-binned boundary montage
```{r montage-cell-reg, fig.height = 3.5}
plot_csp_boundary_montage(
  group_result           = cell_cca_reg,
  tpca_result            = tpca_cell_info,
  score_type = "CSP",
  cc_idx                 = 1L,
  num_bins               = 10L,
  num_cells_for_boundary = 5L,
  norm_rescale           = TRUE
)
```

## CosMx workflow

<details>
<summary><strong>Coming soon: CosMx example</strong></summary>

A complete CosMx example will be added in an upcoming release. It will cover:

- loading Nanostring expression data;
- accessing cell segmentations;
- computing TPCA embeddings;
- running CCA between morphology and expression factors.

</details>
