---
title: "Integrating morphology and gene expression with kstitch"
subtitle: "A Xenium Prime melanoma keratinocyte case study"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integrating morphology and gene expression with kstitch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

Spatial transcriptomics tells us *where* genes are expressed, but the shape of
a cell — how round it is, how large its nucleus, how irregular its boundary —
carries independent biological information that transcript counts alone cannot
capture. **kstitch** links these two modalities by computing a low-dimensional
representation of cell morphology (via Tangent PCA, or TPCA) and finding the
axes of maximal correlation between shape and gene expression (via canonical
correlation analysis, or CCA).

This vignette walks through a complete kstitch analysis on a subset of
keratinocytes from a human melanoma Xenium Prime dataset. The object used here
has been pre-computed (TPCA, NMF, and CCA have already been run and stored
inside the object) so that the vignette can be reproduced without Python or
`rliger`. The script that built this object lives in
`data-raw/prepare_vignette_data.R`.

### What is TPCA?

Standard PCA operates on flat Euclidean space. Cell contour polygons live on a
curved manifold (Kendall's shape space), where straight-line interpolation
between shapes is geometrically meaningless. Tangent PCA (also called Principal
Geodesic Analysis, or PGA) works by:

1. Finding the **Fréchet mean** — the average shape in the curved space.
2. Projecting all shapes onto the **tangent plane** at that mean via a log-map.
3. Running standard SVD on the projected (now flat) coordinates.

The result is a set of **shape principal components** (Shape PCs) that
describe the major axes of morphological variation in a geometrically correct
way.

### What does CCA find?

CCA finds linear combinations of the shape PCs (called **CSP** — Canonical
Shape variates) and linear combinations of the NMF factors (called **CEP** —
Canonical Expression variates) such that each CSP/CEP pair has maximal
correlation. A high canonical correlation on the first component means there is
a strong axis of joint morpho-molecular variation: cells that differ in shape
along CSP1 also differ in gene expression along CEP1.

---

## Setup

```{r load}
library(kstitch)
library(ggplot2)
```

Load the pre-computed vignette object:

```{r load-obj}
vignette_data         <- readRDS(system.file("extdata", "kstitch_vignette_xenium.rds",
                                             package = "kstitch"))
obj                   <- vignette_data$obj
tpca_cell_metadata    <- vignette_data$metadata$tpca_cell
tpca_nucleus_metadata <- vignette_data$metadata$tpca_nucleus
```

The object contains 3,171 keratinocytes. Of these, 2,331 have TPCA shape PCs
computed for both cell and nucleus boundaries. The remaining 840 cells have
`NA` in the TPCA reductions — these are cells that were excluded from the PGA
run (e.g. cells with degenerate or missing boundary polygons).

```{r check-reductions}
names(obj@reductions)
sum(!is.na(Seurat::Embeddings(obj, "tpca_nucleus")[, 1]))
```

---

## Step 1 — Inspect shape modes

The `plot_shape_modes()` function reconstructs and plots cell shapes along each
principal geodesic component at a range of standard deviation values. This is
the most direct way to understand what each shape PC captures biologically.

```{r shape-modes, fig.height = 6}
tpca_nucleus_info <- list(
  Metadata       = tpca_nucleus_metadata,
  TPCA_Embedding = Seurat::Embeddings(obj, "tpca_nucleus"),
  Info           = get_tpca_results(obj, "nucleus"),
  contour_type   = "nucleus",
  output_dir     = NA_character_
)

plot_shape_modes(
  tpca_nucleus_info,
  sds_to_plot = c(-2, -1, 0, 1, 2),
  num_pcs     = 4L
)
```

Reading these plots: each row is a shape PC, each column is a standard
deviation value along that component. PC1 here is dominated by **nuclear
size** (as expected — area variation is re-introduced as `scaled_log_area`
and tends to dominate the first component). PC2 and beyond capture
**shape** variation independent of size.

---

## Step 2 — Covariate regression

### Why regress covariates?

CCA maximises correlation between the shape and expression sides. Without
adjustment, technical covariates — sequencing depth (`nCount_Xenium`) and
segmentation method — can dominate the canonical axes, obscuring biologically
meaningful morpho-molecular relationships. For example, cells with higher
sequencing depth may appear to have different NMF factor scores simply because
more transcripts were captured, not because their gene expression programme
differs.

We address this by regressing technical covariates from both the shape and
expression matrices before running CCA, using `lme4::lmer`. The regression
models differ between the two sides:

- **Shape matrix** (`shape_pcs + segmentation_method`): regress
  `scaled_log_depth + (1 | segmentation_method)`. Area is deliberately
  retained as a predictor column rather than regressed out — nuclear size is
  biologically meaningful and we want CCA to find axes that may be driven in
  part by size.
- **Expression matrix** (NMF factors): regress
  `scaled_log_area + scaled_log_depth + (1 | segmentation_method)`. Area is
  removed from the expression side because differences in factor scores
  attributable purely to cell size are not the signal of interest here.

The vignette presents CCA results **both with and without** covariate
regression so the reader can assess which canonical axes survive adjustment.

### Shape and expression matrices

In both the regressed and unregressed analyses, the shape matrix is constructed
by concatenating the top 10 nucleus shape PCs with scaled log nuclear area as
an additional column. This lets CCA find axes that may be partly driven by
size, while also capturing shape variation independent of size. The expression
matrix is the full NMF factor score matrix.

```{r retrieve-cca-results}
# Unregressed CCA results
nuc_cca_unreg  <- get_kstitch_results(obj, "nuc_cca_unregressed")
cell_cca_unreg <- get_kstitch_results(obj, "cell_cca_unregressed")

# Regressed CCA results
nuc_cca_reg    <- get_kstitch_results(obj, "nuc_cca_regressed")
cell_cca_reg   <- get_kstitch_results(obj, "cell_cca_regressed")
```

---

## Step 3 — Canonical correlations

Comparing canonical correlations before and after regression reveals how much
of the morpho-molecular signal is attributable to technical covariates versus
biology.

```{r cc-barplot, fig.height = 4}
k <- min(length(nuc_cca_unreg$CC_Corr_Coefs),
         length(nuc_cca_reg$CC_Corr_Coefs))

cc_df <- data.frame(
  Component  = rep(paste0("CC", seq_len(k)), 2),
  Correlation = c(nuc_cca_unreg$CC_Corr_Coefs[seq_len(k)],
                  nuc_cca_reg$CC_Corr_Coefs[seq_len(k)]),
  Analysis   = rep(c("Unregressed", "Regressed"), each = k)
)

ggplot(cc_df, aes(x = Component, y = Correlation, fill = Analysis)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("Unregressed" = "steelblue",
                               "Regressed"   = "tomato")) +
  labs(title = "Nucleus CCA — canonical correlations",
       y     = "Canonical correlation") +
  theme_classic()
```

A value above ~0.3–0.4 generally indicates a biologically meaningful
morpho-molecular axis. Canonical correlations that drop substantially after
regression are likely driven by technical covariates; those that persist
reflect genuine morpho-molecular coupling.

---

## Step 4 — CSP loadings

The CSP loading plot shows which shape features load most strongly on each
canonical shape variate. By convention (enforced by `anchor_cca_signs()`),
the sign of each component is set so that the shape feature with the highest
absolute structure correlation loads positively. The anchor feature is shown
in the subtitle.

### Unregressed

```{r csp-loadings-unreg-1}
plot_csp_loadings(nuc_cca_unreg, cc_idx = 1)
```

```{r csp-loadings-unreg-2}
plot_csp_loadings(nuc_cca_unreg, cc_idx = 2)
```

### Regressed

```{r csp-loadings-reg-1}
plot_csp_loadings(nuc_cca_reg, cc_idx = 1)
```

```{r csp-loadings-reg-2}
plot_csp_loadings(nuc_cca_reg, cc_idx = 2)
```

CSP1 in the unregressed analysis is typically dominated by `scaled_log_area`.
After regression, this size signal is attenuated and shape PCs contribute more
to CSP1, revealing morphological structure that was previously masked.

---

## Step 5 — CEP loadings

The CEP loading plot shows which NMF factors are most associated with each
morphological axis. Factors with large positive loadings on CEP1 are
up-regulated in cells with large CSP1 scores.

### Unregressed

```{r cep-loadings-unreg-1}
plot_cep_loadings(nuc_cca_unreg, cc_idx = 1)
```

```{r cep-loadings-unreg-2}
plot_cep_loadings(nuc_cca_unreg, cc_idx = 2)
```

### Regressed

```{r cep-loadings-reg-1}
plot_cep_loadings(nuc_cca_reg, cc_idx = 1)
```

```{r cep-loadings-reg-2}
plot_cep_loadings(nuc_cca_reg, cc_idx = 2)
```

To interpret which genes drive each NMF factor, inspect
`obj@misc$kstitch$nmf$all$NMF_Loading` — each column is a factor, each row
is a gene.

---

## Step 6 — CSP-binned boundary montage

The boundary montage is the headline figure of kstitch. Cells are binned by
their CSP score into deciles. Within each bin, the cells closest to the bin
centre (medoids, measured by pairwise PGA distance) are selected and their
contours are Procrustes-aligned to the Fréchet mean and displayed.

By default (`norm_rescale = TRUE`), contours are rescaled by the contour's
Frobenius norm so that the displayed size reflects true nuclear area — the
size gradient you see across bins is real.

### Unregressed

```{r montage-unreg, fig.height = 3.5}
plot_csp_boundary_montage(
  group_result           = nuc_cca_unreg,
  tpca_result            = tpca_nucleus_info,
  cc_idx                 = 1L,
  num_bins               = 10L,
  num_cells_for_boundary = 5L,
  norm_rescale           = TRUE
)
```

### Regressed

```{r montage-reg, fig.height = 3.5}
plot_csp_boundary_montage(
  group_result           = nuc_cca_reg,
  tpca_result            = tpca_nucleus_info,
  cc_idx                 = 1L,
  num_bins               = 10L,
  num_cells_for_boundary = 5L,
  norm_rescale           = TRUE
)
```

> **Note:** Because `pre_shape_embedding` was not shipped with this vignette
> object (to keep file size small), shapes are reconstructed from the top PCA
> components. Fine morphological detail may differ slightly from the original
> contours — this is noted in the plot subtitle.

---

## Step 7 — Repeat for cell contours

Everything above can be repeated for cell (rather than nucleus) contours by
substituting `cell_cca_unregressed` and `cell_cca_regressed`:

```{r cell-cca}
plot_csp_loadings(cell_cca_unreg, cc_idx = 1)
plot_csp_loadings(cell_cca_reg,   cc_idx = 1)
plot_cep_loadings(cell_cca_unreg, cc_idx = 1)
plot_cep_loadings(cell_cca_reg,   cc_idx = 1)
```

Comparing nucleus and cell CCA results reveals which morpho-molecular axes are
driven by nuclear shape specifically versus whole-cell shape. Comparing
regressed and unregressed results reveals which axes reflect biology versus
technical variation.

---

## Running the full pipeline on your own data

The vignette above skips the TPCA and NMF computation steps to keep it
self-contained. For a full analysis on your own Xenium or CosMx data:

```{r full-pipeline, eval = FALSE}
# seg_list is the list returned by ReadXenium() or ReadNanostring(), e.g.:
# seg_list <- ReadXenium("path/to/xenium/output", type = c("cell", "nucleus"))

# 1. Run TPCA (starting from a Seurat object with segmentation list)
tpca_result <- export_seurat_contours(
  obj          = obj,
  seg_list     = seg_list,
  contour_type = "nucleus",
  output_dir   = "output/tpca_nucleus"
)
obj <- store_tpca_results(obj, tpca_result)

# 2. Run NMF
nmf_results <- compute_nmf(
  obj        = obj,
  assay_name = "Xenium",
  group.by   = "cell_type"
)
obj <- store_nmf_results(obj, nmf_results)
```

### Covariate regression

Before running CCA, we recommend regressing technical covariates from both the
shape and expression matrices. The shape matrix is constructed by concatenating
the top shape PCs with scaled log area; the expression matrix is the NMF factor
score matrix. Both are then adjusted for sequencing depth and segmentation
method using `lme4::lmer`:

```{r covariate-regression, eval = FALSE}
library(lme4)

shape_pcs <- Seurat::Embeddings(obj, "tpca_nucleus")[, 1:10]
expr_mat  <- Seurat::Embeddings(obj, "nmf_all")
meta      <- obj@meta.data

# Build covariate vectors, aligned to shape matrix row order
area_vec  <- setNames(log(meta[rownames(shape_pcs), "nucleus_area"]),
                      rownames(shape_pcs))
depth_vec <- setNames(log(meta[rownames(shape_pcs), "nCount_Xenium"]),
                      rownames(shape_pcs))
seg_vec   <- setNames(meta[rownames(shape_pcs), "segmentation_method"],
                      rownames(shape_pcs))

scaled_log_area  <- scale(area_vec)[, 1]
scaled_log_depth <- scale(depth_vec)[, 1]

# Concatenate scaled log area onto shape PCs
shape_mat <- cbind(shape_pcs, scaled_log_area = scaled_log_area)

# Regression helper: fits lmer per column, returns residuals
regress_lmer <- function(mat, log_area, log_depth, seg_method) {
  cell_ids  <- rownames(mat)
  resid_mat <- matrix(NA_real_, nrow = nrow(mat), ncol = ncol(mat),
                      dimnames = dimnames(mat))
  for (j in seq_len(ncol(mat))) {
    df  <- data.frame(y = mat[, j],
                      log_area  = log_area[cell_ids],
                      log_depth = log_depth[cell_ids],
                      seg       = seg_method[cell_ids])
    ok  <- complete.cases(df)
    fit <- lmer(y ~ log_area + log_depth + (1 | seg), data = df[ok, ],
                REML = FALSE)
    resid_mat[ok, j] <- residuals(fit)
  }
  resid_mat
}

shape_mat_reg <- regress_lmer(shape_mat, scaled_log_area, scaled_log_depth, seg_vec)
expr_mat_reg  <- regress_lmer(expr_mat,  scaled_log_area, scaled_log_depth, seg_vec)
```

### Running CCA with and without regression

```{r run-cca, eval = FALSE}
# Unregressed
cca_unreg <- link_shape_and_factors(
  obj       = obj,
  shape_mat = shape_mat,
  expr_mat  = expr_mat,
  group.by  = NULL
)
obj@misc$kstitch$all$nucleus_cca_unregressed <- anchor_cca_signs(cca_unreg$all)

# Regressed
cca_reg <- link_shape_and_factors(
  obj       = obj,
  shape_mat = shape_mat_reg,
  expr_mat  = expr_mat_reg,
  group.by  = NULL
)
obj@misc$kstitch$all$nucleus_cca_regressed <- anchor_cca_signs(cca_reg$all)
```

See `?export_seurat_contours`, `?compute_nmf`, and `?link_shape_and_factors`
for full argument documentation.
