---
title: "Integrating morphology and gene expression with kstitch"
subtitle: "A Xenium Prime melanoma keratinocyte case study"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integrating morphology and gene expression with kstitch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

Spatial transcriptomics tells us *where* genes are expressed, but the shape of
a cell — how round it is, how large its nucleus, how irregular its boundary —
carries independent biological information that transcript counts alone cannot
capture. **kstitch** links these two modalities by computing a low-dimensional
representation of cell morphology (via Tangent PCA, or TPCA) and finding the
axes of maximal correlation between shape and gene expression (via canonical
correlation analysis, or CCA).

This vignette walks through a complete kstitch analysis on a subset of
keratinocytes from a human melanoma Xenium Prime dataset. The object used here
has been pre-computed (TPCA, NMF, and CCA have already been run and stored
inside the object) so that the vignette can be reproduced without Python or
`rliger`. The script that built this object lives in
`data-raw/prepare_vignette_data.R`.

### What is TPCA?

Standard PCA operates on flat Euclidean space. Cell contour polygons live on a
curved manifold (Kendall's shape space), where straight-line interpolation
between shapes is geometrically meaningless. Tangent PCA (also called Principal
Geodesic Analysis, or PGA) works by:

1. Finding the **Fréchet mean** — the average shape in the curved space.
2. Projecting all shapes onto the **tangent plane** at that mean via a log-map.
3. Running standard SVD on the projected (now flat) coordinates.

The result is a set of **shape principal components** (Shape PCs) that
describe the major axes of morphological variation in a geometrically correct
way.

### What does CCA find?

CCA finds linear combinations of the shape PCs (called **CSP** — Canonical
Shape variates) and linear combinations of the NMF factors (called **CEP** —
Canonical Expression variates) such that each CSP/CEP pair has maximal
correlation. A high canonical correlation on the first component means there is
a strong axis of joint morpho-molecular variation: cells that differ in shape
along CSP1 also differ in gene expression along CEP1.

---

## Setup

```{r load}
library(kstitch)
library(ggplot2)
```

Load the pre-computed vignette object:

```{r load-obj}
obj <- readRDS(system.file("extdata", "kstitch_vignette_xenium.rds",
                           package = "kstitch"))
obj
```

The object contains 3,171 keratinocytes. Of these, 2,331 have TPCA shape PCs
computed for both cell and nucleus boundaries. The remaining 840 cells have
`NA` in the TPCA reductions — these are cells that were excluded from the PGA
run (e.g. cells with degenerate or missing boundary polygons).

```{r check-reductions}
# Available reductions
names(obj@reductions)

# How many cells have TPCA embeddings
sum(!is.na(Seurat::Embeddings(obj, "tpca_nucleus")[, 1]))
```

---

## Step 1 — Inspect shape modes

The `plot_shape_modes()` function reconstructs and plots cell shapes along each
principal geodesic component at a range of standard deviation values. This is
the most direct way to understand what each shape PC captures biologically.

```{r shape-modes, fig.height = 6}
tpca_nucleus_info <- list(
  TPCA_Embedding = Seurat::Embeddings(obj, "tpca_nucleus"),
  Info          = get_tpca_results(obj, "nucleus"),
  contour_type  = "nucleus",
  output_dir    = NA_character_
)

plot_shape_modes(
  tpca_nucleus_info,
  sds_to_plot = c(-2, -1, 0, 1, 2),
  num_pcs     = 4L
)
```

Reading these plots: each row is a shape PC, each column is a standard
deviation value along that component. PC1 here is dominated by **nuclear
size** (as expected — area variation is re-introduced as `scaled_log_area`
and tends to dominate the first component). PC2 and beyond capture
**shape** variation independent of size.

---

## Step 2 — Inspect canonical correlations

Retrieve the CCA results for the nucleus contour analysis:

```{r get-cca}
nuc_cca <- obj@misc$kstitch$nucleus_cca

# Canonical correlations across components
barplot(nuc_cca$CC_Corr_Coefs,
        names.arg = paste0("CC", seq_along(nuc_cca$CC_Corr_Coefs)),
        ylab = "Canonical correlation",
        main = "Nucleus CCA — canonical correlations",
        col  = "steelblue")
```

The first few canonical correlations are typically the most interpretable.
A value above ~0.3–0.4 generally indicates a biologically meaningful
morpho-molecular axis.

---

## Step 3 — CSP loadings

The CSP loading plot shows which shape features load most strongly on each
canonical shape variate. By convention (enforced by `anchor_cca_signs()`),
the sign of each component is set so that the shape feature with the highest
absolute structure correlation loads **positively**. The anchor feature is
shown in the subtitle.

```{r csp-loadings}
plot_csp_loadings(nuc_cca, cc_idx = 1)
```

```{r csp-loadings-2}
plot_csp_loadings(nuc_cca, cc_idx = 2)
```

CSP1 is typically dominated by nuclear area (large positive loading on
`scaled_log_area`). CSP2 and beyond capture shape variation independent of
size — look for loadings on Shape PC2, PC3, etc.

---

## Step 4 — CEP loadings

The CEP loading plot shows which NMF factors are most associated with each
morphological axis. Factors with large positive loadings on CEP1 are
up-regulated in cells with large CSP1 scores (i.e. cells with the morphology
that CSP1 represents).

```{r cep-loadings}
plot_cep_loadings(nuc_cca, cc_idx = 1)
```

```{r cep-loadings-2}
plot_cep_loadings(nuc_cca, cc_idx = 2)
```

To interpret which genes drive each NMF factor, inspect
`obj@misc$kstitch$nmf$NMF_Loading` — each column is a factor, each row
is a gene.

---

## Step 5 — CSP-binned boundary montage

The boundary montage is the headline figure of kstitch. Cells are binned by
their CSP score into deciles. Within each bin, the cells closest to the bin
centre (medoids, measured by pairwise PGA distance) are selected and their
contours are Procrustes-aligned to the Fréchet mean and displayed.

By default (`area_rescale = TRUE`), contours are rescaled by `sqrt(area)` so
that the displayed size reflects true nuclear area — the size gradient you see
across bins is real.

# Note: area_rescale = TRUE is requested here, but this vignette's
# tpca_nucleus_info sets output_dir = NA_character_ (no Shape_Metadata.csv.gz
# is shipped with the vignette object). plot_csp_boundary_montage() checks
# file.exists(file.path(output_dir, "Shape_Metadata.csv.gz")) before
# rescaling, so area_rescale silently falls back to FALSE here -- contours
# below are NOT area-rescaled despite the argument. On your own data, with a
# real output_dir containing Shape_Metadata.csv.gz, area_rescale = TRUE will
# work as documented.

```{r montage, fig.height = 3.5}
plot_csp_boundary_montage(
  group_result           = nuc_cca,
  tpca_result            = tpca_nucleus_info,
  cc_idx                 = 1L,
  num_bins               = 10L,
  num_cells_for_boundary = 5L,
  area_rescale           = TRUE
)
```

> **Note:** Because `pre_shape_embedding` was not shipped with this vignette
> object (to keep file size small), shapes are reconstructed from the top PCA
> components. Fine morphological detail may differ slightly from the original
> contours — this is noted in the plot subtitle.

---

## Step 6 — Repeat for cell contours

Everything above can be repeated for cell (rather than nucleus) contours by
substituting `cell_cca` and `tpca_cell`:

```{r cell-cca, eval = TRUE}
cell_cca <- obj@misc$kstitch$cell_cca

plot_csp_loadings(cell_cca, cc_idx = 1)
plot_cep_loadings(cell_cca, cc_idx = 1)
```

Comparing nucleus and cell CCA results reveals which morpho-molecular axes are
driven by nuclear shape specifically versus whole-cell shape.

---

## Running the full pipeline on your own data

The vignette above skips the TPCA and NMF computation steps to keep it
self-contained. For a full analysis on your own Xenium or CosMx data:

```r
# 1. Export boundary parquet (done externally — see data-raw/prepare_vignette_data.R)

# 2. Run TPCA
tpca_result <- run_tpca(
  boundary_parquet_path = "nucleus_boundaries.parquet",
  output_dir            = "output/tpca_nucleus",
  contour_type          = "nucleus"
)
obj <- store_tpca_results(obj, tpca_result)

# 3. Run NMF
nmf_results <- compute_nmf(
  obj        = obj,
  assay_name = "Xenium",
  group.by   = "cell_type"
)
## Running the full pipeline on your own data

The vignette above skips the TPCA and NMF computation steps to keep it
self-contained. For a full analysis on your own Xenium or CosMx data:

```r
# 1. Export boundary parquet (done externally — see data-raw/prepare_vignette_data.R)

# 2. Run TPCA
tpca_result <- run_tpca(
  boundary_parquet_path = "nucleus_boundaries.parquet",
  output_dir            = "output/tpca_nucleus",
  contour_type          = "nucleus"
)
obj <- store_tpca_results(obj, tpca_result)

# 3. Run NMF
nmf_results <- compute_nmf(
  obj        = obj,
  assay_name = "Xenium",
  group.by   = "cell_type"
)
```

### Concatenating covariates onto the shape matrix

In the manuscript accompanying this package, CCA is run between gene
expression and a shape matrix that has cell (or nucleus) **area concatenated
on as an extra column**, rather than treating area as a nuisance covariate to
regress out. This lets CCA find canonical axes that may be driven in part by
size, which is often biologically meaningful (e.g. larger nuclei tend to
correlate with specific transcriptional states).

Area is typically already present in `obj@meta.data` for Xenium and CosMx
objects (check `colnames(obj@meta.data)` for the exact column name in your
object — it may be `cell_area`, `nucleus_area`, or similar depending on how
the object was loaded).

The important part is **alignment**: `area_vec` must be in the exact same
cell order as the shape matrix before you `cbind()` them together, or the
area values will silently get attached to the wrong cells. The safe pattern
is to always index the covariate *by the shape matrix's own row names*:

```r
shape_pcs <- Seurat::Embeddings(obj, "tpca_nucleus")[, 1:10]

# Pull area from meta.data, explicitly aligned to shape_pcs's cell order.
# Indexing by rownames(shape_pcs) first guarantees the result is already in
# the right order; re-attaching the same names makes that explicit.
area_vec <- setNames(
  obj@meta.data[rownames(shape_pcs), "cell_area"],
  rownames(shape_pcs)
)

# Concatenate as an extra column -- safe now that both share the same row
# order by construction. Log-transforming area is common practice since
# cell/nucleus area is typically right-skewed.
shape_mat <- cbind(
  shape_pcs,
  scaled_log_area = scale(log(area_vec))[, 1]
)
```

The same pattern applies to concatenating covariates onto the **expression**
side (`expr_mat`) rather than the shape side — for example, sequencing depth
(`nCount_RNA`, already present in every Seurat object's `@meta.data` by
default):

```r
expr_mat <- nmf_results$YourCellType$NMF_Matrix

depth_vec <- setNames(
  obj@meta.data[rownames(expr_mat), "nCount_RNA"],
  rownames(expr_mat)
)

expr_mat <- cbind(expr_mat, scaled_log_depth = scale(log(depth_vec))[, 1])
```

> **CosMx note:** for CosMx data, a natural additional covariate is each
> cell's distance to the nearest field-of-view (FOV) edge, since
> segmentation quality tends to degrade near FOV boundaries. Computing this
> distance requires FOV polygon/tile boundary information, which is not
> always a direct `meta.data` column — see your loading tool's
> documentation (e.g. Seurat's `LoadNanostring()`/`ReadNanostring()`, or the
> `spacetrooper` package) for how FOV geometry is exposed for your object.
> This will be covered in a future revision of this vignette.

### Running CCA with the concatenated matrix

```r
cca_results <- link_shape_and_factors(
  obj       = obj,
  expr_mat  = expr_mat,
  shape_mat = shape_mat,
  group.by  = "cell_type"
)
obj <- store_kstitch_results(obj, cca_results)
```

See `?run_tpca`, `?compute_nmf`, and `?link_shape_and_factors` for full
argument documentation.
