---
title: "Integrating morphology and gene expression with kstitch"
subtitle: "A Xenium Prime melanoma keratinocyte case study"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Integrating morphology and gene expression with kstitch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 5,
  out.width = "100%"
)
```

## Introduction

Spatial transcriptomics tells us *where* genes are expressed, but the shape of
a cell — how round it is, how large its nucleus, how irregular its boundary —
carries independent biological information that transcript counts alone cannot
capture. **kstitch** links these two modalities by computing a low-dimensional
representation of cell morphology (via Tangent PCA, or TPCA) and finding the
axes of maximal correlation between shape and gene expression (via canonical
correlation analysis, or CCA).

This vignette walks through a complete kstitch analysis on a subset of
keratinocytes from a human melanoma Xenium Prime dataset. The object used here
has been pre-computed (TPCA, NMF, and CCA have already been run and stored
inside the object) so that the vignette can be reproduced without Python or
`rliger`. The script that built this object lives in
`data-raw/prepare_vignette_data.R`.

### What is TPCA?

Standard PCA operates on flat Euclidean space. Cell contour polygons live on a
curved manifold (Kendall's shape space), where straight-line interpolation
between shapes is geometrically meaningless. Tangent PCA (also called Principal
Geodesic Analysis, or PGA) works by:

1. Finding the **Fréchet mean** — the average shape in the curved space.
2. Projecting all shapes onto the **tangent plane** at that mean via a log-map.
3. Running standard SVD on the projected (now flat) coordinates.

The result is a set of **shape principal components** (Shape PCs) that describe
the major axes of morphological variation in a geometrically correct way.

### What does CCA find?

CCA finds linear combinations of the shape PCs (called **CSP** — Canonical
Shape variates) and linear combinations of the NMF factors (called **CEP** —
Canonical Expression variates) such that each CSP/CEP pair has maximal
correlation. A high canonical correlation on the first component means there is
a strong axis of joint morpho-molecular variation: cells that differ in shape
along CSP1 also differ in gene expression along CEP1.

---

## Setup

```{r load-libs}
library(kstitch)
library(ggplot2)
```

Load the pre-computed vignette object. If you ran the full pipeline chunks
below, `keratinocyte_obj` will already be in memory and you can skip this chunk.

```{r load-obj}
vignette_data         <- readRDS(system.file("extdata", "kstitch_vignette_xenium.rds",
                                             package = "kstitch"))
keratinocyte_obj      <- vignette_data$obj
tpca_nucleus_metadata <- vignette_data$metadata$tpca_nucleus
tpca_cell_metadata    <- vignette_data$metadata$tpca_cell
```

The object contains 3,171 keratinocytes. 2,331 have nucleus TPCA shape PCs
computed and 2,831 have cell TPCA shape PCs computed. Cells missing from each
reduction had degenerate or missing boundary polygons and were excluded from
the PGA run.

```{r check-reductions}
names(keratinocyte_obj@reductions)
sum(!is.na(Seurat::Embeddings(keratinocyte_obj, "tpca_nucleus")[, 1]))
sum(!is.na(Seurat::Embeddings(keratinocyte_obj, "tpca_cell")[, 1]))
```

---

## Step 1 — Download data and run NMF (eval=FALSE)

NMF is run once and its factors feed into both the nuclear and cell shape
analyses below.

### Download the data

```{r download-data, eval = FALSE}
# Download the Xenium Prime Human Skin FFPE dataset from 10x Genomics and
# unpack it into a local directory. Update xenium_data_path to match.
xenium_data_path <- "#path/to/Xenium_Prime_Human_Skin_FFPE_outs/"
```

### Load the Xenium object and subset to keratinocytes

```{r subset-keratinocytes, eval = FALSE}
library(Seurat)
library(kstitch)

vignette_data <- readRDS(system.file("extdata", "kstitch_vignette_xenium.rds",
                                     package = "kstitch"))
precomputed_obj <- vignette_data$obj

keratinocyte_cell_ids     <- Cells(precomputed_obj)
keratinocyte_genes_to_use <- Features(precomputed_obj)  # genes >= 10 nTPM in
                                                         # keratinocytes (HPA)

melanoma_xenium_obj <- LoadXenium(xenium_data_path, molecule.coordinates = FALSE)
keratinocyte_obj <- subset(melanoma_xenium_obj,
                           cells    = keratinocyte_cell_ids,
                           features = keratinocyte_genes_to_use)
# You may receive object-validation warnings; these can be ignored.
```

### Run NMF

```{r run-nmf, eval = FALSE}
nmf_output <- compute_nmf(
  keratinocyte_obj,
  assay_name               = "Xenium",
  stability_seed           = 235,
  default_num_factors      = 10,
  num_top_genes_per_factor = 50,
  integration_method       = "online",
  selectGenes_params       = list(nGenes = "all", verbose = TRUE),
  gene_pct_threshold       = 0.01
)

keratinocyte_obj <- store_nmf_results(keratinocyte_obj, nmf_output)
```

### Load segmentation info

```{r load-seg, eval = FALSE}
melanoma_xenium_segmentation_info <- ReadXenium(
  xenium_data_path,
  outs = c("segmentation_method"),
  type = c("segmentations", "nucleus_segmentations", "centroids")
)
```

### Covariate regression helper

We define `.regress_lmer()` once here for use in both the nuclear and cell
shape analyses. It operates column-by-column on a matrix: for each column it
fits a mixed-effects model with `scaled_log_depth` as a fixed effect and
`segmentation_method` as a random intercept, then returns the model residuals.
Cells with incomplete covariate data are excluded from model fitting and left
as `NA` in the output.

```{r define-regress-lmer, eval = FALSE}
library(lme4)

.regress_lmer <- function(mat, scaled_log_depth, seg_method) {
  cell_ids  <- rownames(mat)
  resid_mat <- matrix(NA_real_, nrow = nrow(mat), ncol = ncol(mat),
                      dimnames = dimnames(mat))
  for (j in seq_len(ncol(mat))) {
    df <- data.frame(
      y                = mat[, j],
      scaled_log_depth = scaled_log_depth[cell_ids],
      seg_method       = seg_method[cell_ids],
      stringsAsFactors = FALSE
    )
    ok <- complete.cases(df)
    if (sum(ok) < 10L) {
      warning(sprintf("Column %d: fewer than 10 complete cases, skipping.", j))
      next
    }
    fit <- lmer(y ~ scaled_log_depth + (1 | seg_method),
                data = df[ok, ], REML = FALSE)
    resid_mat[ok, j] <- residuals(fit)
  }
  resid_mat
}

meta           <- keratinocyte_obj@meta.data
log_depth_vec  <- setNames(log(meta$nCount_Xenium),  rownames(meta))
seg_method_vec <- setNames(meta$segmentation_method, rownames(meta))
nmf_cells      <- rownames(Embeddings(keratinocyte_obj, "nmf_all"))
```

---

# Nuclear shape analysis

## Step 2 — Export nucleus contours and run TPCA (eval=FALSE)

TPCA is restricted to cells that have exactly one nucleus.

```{r run-tpca-nucleus, eval = FALSE}
keratinocytes_cells_with_single_nucleus <-
  rownames(na.omit(Embeddings(precomputed_obj, "tpca_nucleus")))

tpca_nucleus_out <- export_seurat_contours(
  melanoma_xenium_obj,
  seg_list     = melanoma_xenium_segmentation_info,
  output_dir   = "output/tpca_nucleus",
  contour_type = "nucleus",
  use_parallel = TRUE,
  num_threads  = 8,
  cell_ids     = keratinocytes_cells_with_single_nucleus
)
keratinocyte_obj <- store_tpca_results(keratinocyte_obj, tpca_nucleus_out,
                                       reduction_key_prefix = "NucleusShapePC_")
```

## Step 3 — Nucleus covariate regression (eval=FALSE)

The shape matrix is the top 10 nucleus shape PCs concatenated with scaled log
nuclear area. Scaled log area is retained as a predictor column rather than
regressed out — nuclear size is biologically meaningful and we want CCA to
find axes that may be partly driven by it. Scaled log sequencing depth and
segmentation method are regressed from both the shape and expression matrices.

```{r nucleus-regression, eval = FALSE}
tpca_nuc_cells   <- rownames(Embeddings(keratinocyte_obj, "tpca_nucleus"))
shared_nuc_cells <- intersect(tpca_nuc_cells, nmf_cells)

nuc_log_area_vec     <- setNames(
  log(tpca_nucleus_out$all$Metadata$area[
        match(rownames(meta), tpca_nucleus_out$all$Metadata$cell_id)]),
  rownames(meta)
)
scaled_nuc_log_area  <- setNames(scale(nuc_log_area_vec[shared_nuc_cells])[, 1],
                                  shared_nuc_cells)
scaled_log_depth_nuc <- setNames(scale(log_depth_vec[shared_nuc_cells])[, 1],
                                  shared_nuc_cells)

shape_pcs_nuc       <- Embeddings(keratinocyte_obj, "tpca_nucleus")[shared_nuc_cells, 1:10]
expr_mat_nuc        <- Embeddings(keratinocyte_obj, "nmf_all")[shared_nuc_cells, ]
shape_mat_nuc_unreg <- cbind(shape_pcs_nuc,
                              scaled_log_area = scaled_nuc_log_area[shared_nuc_cells])

shape_mat_nuc_reg <- .regress_lmer(shape_mat_nuc_unreg,
                                    scaled_log_depth_nuc, seg_method_vec)
expr_mat_nuc_reg  <- .regress_lmer(expr_mat_nuc,
                                    scaled_log_depth_nuc, seg_method_vec)
```

## Step 4 — Nucleus CCA (eval=FALSE)

```{r nucleus-cca, eval = FALSE}
cca_nuc_unreg <- link_shape_and_factors(
  obj       = keratinocyte_obj,
  shape_mat = shape_mat_nuc_unreg,
  expr_mat  = expr_mat_nuc,
  group.by  = NULL
)
keratinocyte_obj <- store_kstitch_results(keratinocyte_obj, cca_nuc_unreg,
                                           reduction_name    = "nuc_cca_unregressed",
                                           reduction_key_csp = "CSPNU_",
                                           reduction_key_cep = "CEPNU_")

cca_nuc_reg <- link_shape_and_factors(
  obj       = keratinocyte_obj,
  shape_mat = shape_mat_nuc_reg,
  expr_mat  = expr_mat_nuc_reg,
  group.by  = NULL
)
keratinocyte_obj <- store_kstitch_results(keratinocyte_obj, cca_nuc_reg,
                                           reduction_name    = "nuc_cca_regressed",
                                           reduction_key_csp = "CSPNR_",
                                           reduction_key_cep = "CEPNR_")
```

## Nuclear shape results

Retrieve the nucleus CCA results and build the `tpca_result` object expected
by the figure functions. If you ran the full pipeline, use
`tpca_nucleus_out$all$Metadata` in place of `tpca_nucleus_metadata`.

```{r retrieve-nuc-cca}
nuc_cca_unreg <- get_kstitch_results(keratinocyte_obj, "nuc_cca_unregressed")
nuc_cca_reg   <- get_kstitch_results(keratinocyte_obj, "nuc_cca_regressed")

tpca_nucleus_info <- list(
  Metadata       = tpca_nucleus_metadata,
  TPCA_Embedding = Seurat::Embeddings(keratinocyte_obj, "tpca_nucleus"),
  Info           = get_tpca_results(keratinocyte_obj, "nucleus"),
  contour_type   = "nucleus",
  output_dir     = NA_character_
)
```

### Shape modes

`plot_shape_modes()` reconstructs nucleus shapes along each principal geodesic
component at a range of standard deviation values — the most direct way to
understand what each shape PC captures biologically. Each row is a shape PC;
each column is a standard deviation along that component. PC1 is typically
dominated by **nuclear size** (area variation enters as `scaled_log_area` and
tends to dominate the first component). PC2 and beyond capture shape variation
independent of size.

```{r shape-modes-nuc, fig.height = 6}
plot_shape_modes(
  tpca_nucleus_info,
  sds_to_plot = c(-2, -1, 0, 1, 2),
  num_pcs     = 4L
)
```

### Canonical correlations

Comparing canonical correlations before and after covariate regression reveals
how much of the morpho-molecular signal is attributable to technical covariates
versus biology. Correlations that drop substantially after regression are likely
covariate-driven; those that persist reflect genuine morpho-molecular coupling.
A value above ~0.3–0.4 generally indicates a biologically meaningful axis.

```{r cc-barplot-nuc, fig.height = 4}
k <- min(length(nuc_cca_unreg$all$CC_Corr_Coefs),
         length(nuc_cca_reg$all$CC_Corr_Coefs))

cc_df <- data.frame(
  Component   = rep(paste0("CC", seq_len(k)), 2),
  Correlation = c(nuc_cca_unreg$all$CC_Corr_Coefs[seq_len(k)],
                  nuc_cca_reg$all$CC_Corr_Coefs[seq_len(k)]),
  Analysis    = rep(c("Unregressed", "Regressed"), each = k)
)

ggplot(cc_df, aes(x = Component, y = Correlation, fill = Analysis)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("Unregressed" = "steelblue",
                               "Regressed"   = "tomato")) +
  labs(title = "Nucleus CCA — canonical correlations",
       y     = "Canonical correlation") +
  theme_classic()
```

### CSP loadings

The CSP loading plot shows which shape features load most strongly on each
canonical shape variate. By convention (enforced by `anchor_cca_signs()`),
the sign of each component is set so that the shape feature with the highest
absolute structure correlation loads positively. CSP1 in the unregressed
analysis is typically dominated by `scaled_log_area`. After regression, this
size signal is attenuated and shape PCs contribute more to CSP1.

#### Unregressed

```{r csp-loadings-nuc-unreg-1}
plot_csp_loadings(nuc_cca_unreg$all, cc_idx = 1)
```

```{r csp-loadings-nuc-unreg-2}
plot_csp_loadings(nuc_cca_unreg$all, cc_idx = 2)
```

#### Regressed

```{r csp-loadings-nuc-reg-1}
plot_csp_loadings(nuc_cca_reg$all, cc_idx = 1)
```

```{r csp-loadings-nuc-reg-2}
plot_csp_loadings(nuc_cca_reg$all, cc_idx = 2)
```

### CEP loadings

The CEP loading plot shows which NMF factors are most associated with each
morphological axis. Factors with large positive loadings on CEP1 are
up-regulated in cells with high CSP1 scores. To identify the genes driving
each NMF factor, inspect `keratinocyte_obj@misc$kstitch$nmf$all$NMF_Loading`.

#### Unregressed

```{r cep-loadings-nuc-unreg-1}
plot_cep_loadings(nuc_cca_unreg$all, cc_idx = 1)
```

```{r cep-loadings-nuc-unreg-2}
plot_cep_loadings(nuc_cca_unreg$all, cc_idx = 2)
```

#### Regressed

```{r cep-loadings-nuc-reg-1}
plot_cep_loadings(nuc_cca_reg$all, cc_idx = 1)
```

```{r cep-loadings-nuc-reg-2}
plot_cep_loadings(nuc_cca_reg$all, cc_idx = 2)
```

### CSP-binned boundary montage

The boundary montage is the headline figure of kstitch. Cells are binned by
their CSP score into deciles. Within each bin the cells closest to the bin
centre (medoids, measured by pairwise PGA distance) are selected, their
contours are Procrustes-aligned to the Fréchet mean, and displayed. With
`norm_rescale = TRUE`, contours are rescaled by the contour's Frobenius norm
so that the displayed size reflects true nuclear area.

> **Note:** Because `pre_shape_embedding` was not shipped with this vignette
> object (to keep file size small), shapes are reconstructed from the top PCA
> components. Fine morphological detail may differ slightly from the original
> contours — this is noted in the plot subtitle.

#### Unregressed

```{r montage-nuc-unreg, fig.height = 3.5}
plot_csp_boundary_montage(
  group_result           = nuc_cca_unreg$all,
  tpca_result            = tpca_nucleus_info,
  cc_idx                 = 1L,
  num_bins               = 10L,
  num_cells_for_boundary = 5L,
  norm_rescale           = TRUE
)
```

#### Regressed

```{r montage-nuc-reg, fig.height = 3.5}
plot_csp_boundary_montage(
  group_result           = nuc_cca_reg$all,
  tpca_result            = tpca_nucleus_info,
  cc_idx                 = 1L,
  num_bins               = 10L,
  num_cells_for_boundary = 5L,
  norm_rescale           = TRUE
)
```

---

# Cell shape analysis

The cell shape analysis follows exactly the same structure as the nuclear shape
analysis above. The key differences are that contours are drawn from the whole
cell boundary rather than the nucleus, all keratinocytes are included (not just
those with a single nucleus), and the CCA results are stored under
`cell_cca_unregressed` and `cell_cca_regressed`. Comparing nucleus and cell
CCA results can reveal whether a morpho-molecular axis is driven by nuclear
shape specifically or by whole-cell shape.

## Step 2 — Export cell contours and run TPCA (eval=FALSE)

```{r run-tpca-cell, eval = FALSE}
tpca_cell_out <- export_seurat_contours(
  melanoma_xenium_obj,
  seg_list     = melanoma_xenium_segmentation_info,
  output_dir   = "output/tpca_cell",
  contour_type = "cell",
  use_parallel = TRUE,
  num_threads  = 8,
  cell_ids     = Cells(keratinocyte_obj)
)
keratinocyte_obj <- store_tpca_results(keratinocyte_obj, tpca_cell_out,
                                       reduction_key_prefix = "CellShapePC_")
```

## Step 3 — Cell covariate regression (eval=FALSE)

```{r cell-regression, eval = FALSE}
tpca_cell_cells   <- rownames(Embeddings(keratinocyte_obj, "tpca_cell"))
shared_cell_cells <- intersect(tpca_cell_cells, nmf_cells)

cell_log_area_vec     <- setNames(
  log(tpca_cell_out$all$Metadata$area[
        match(rownames(meta), tpca_cell_out$all$Metadata$cell_id)]),
  rownames(meta)
)
scaled_cell_log_area  <- setNames(scale(cell_log_area_vec[shared_cell_cells])[, 1],
                                   shared_cell_cells)
scaled_log_depth_cell <- setNames(scale(log_depth_vec[shared_cell_cells])[, 1],
                                   shared_cell_cells)

shape_pcs_cell       <- Embeddings(keratinocyte_obj, "tpca_cell")[shared_cell_cells, 1:10]
expr_mat_cell        <- Embeddings(keratinocyte_obj, "nmf_all")[shared_cell_cells, ]
shape_mat_cell_unreg <- cbind(shape_pcs_cell,
                               scaled_log_area = scaled_cell_log_area[shared_cell_cells])

shape_mat_cell_reg <- .regress_lmer(shape_mat_cell_unreg,
                                     scaled_log_depth_cell, seg_method_vec)
expr_mat_cell_reg  <- .regress_lmer(expr_mat_cell,
                                     scaled_log_depth_cell, seg_method_vec)
```

## Step 4 — Cell CCA (eval=FALSE)

```{r cell-cca, eval = FALSE}
cca_cell_unreg <- link_shape_and_factors(
  obj       = keratinocyte_obj,
  shape_mat = shape_mat_cell_unreg,
  expr_mat  = expr_mat_cell,
  group.by  = NULL
)
keratinocyte_obj <- store_kstitch_results(keratinocyte_obj, cca_cell_unreg,
                                           reduction_name    = "cell_cca_unregressed",
                                           reduction_key_csp = "CSPCU_",
                                           reduction_key_cep = "CEPCU_")

cca_cell_reg <- link_shape_and_factors(
  obj       = keratinocyte_obj,
  shape_mat = shape_mat_cell_reg,
  expr_mat  = expr_mat_cell_reg,
  group.by  = NULL
)
keratinocyte_obj <- store_kstitch_results(keratinocyte_obj, cca_cell_reg,
                                           reduction_name    = "cell_cca_regressed",
                                           reduction_key_csp = "CSPCR_",
                                           reduction_key_cep = "CEPCR_")
```

## Cell shape results

Retrieve the cell CCA results and build the `tpca_result` object. If you ran
the full pipeline, use `tpca_cell_out$all$Metadata` in place of
`tpca_cell_metadata`.

```{r retrieve-cell-cca}
cell_cca_unreg <- get_kstitch_results(keratinocyte_obj, "cell_cca_unregressed")
cell_cca_reg   <- get_kstitch_results(keratinocyte_obj, "cell_cca_regressed")

tpca_cell_info <- list(
  Metadata       = tpca_cell_metadata,
  TPCA_Embedding = Seurat::Embeddings(keratinocyte_obj, "tpca_cell"),
  Info           = get_tpca_results(keratinocyte_obj, "cell"),
  contour_type   = "cell",
  output_dir     = NA_character_
)
```

### Shape modes

```{r shape-modes-cell, fig.height = 6}
plot_shape_modes(
  tpca_cell_info,
  sds_to_plot = c(-2, -1, 0, 1, 2),
  num_pcs     = 4L
)
```

### Canonical correlations

```{r cc-barplot-cell, fig.height = 4}
k <- min(length(cell_cca_unreg$all$CC_Corr_Coefs),
         length(cell_cca_reg$all$CC_Corr_Coefs))

cc_df <- data.frame(
  Component   = rep(paste0("CC", seq_len(k)), 2),
  Correlation = c(cell_cca_unreg$all$CC_Corr_Coefs[seq_len(k)],
                  cell_cca_reg$all$CC_Corr_Coefs[seq_len(k)]),
  Analysis    = rep(c("Unregressed", "Regressed"), each = k)
)

ggplot(cc_df, aes(x = Component, y = Correlation, fill = Analysis)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("Unregressed" = "steelblue",
                               "Regressed"   = "tomato")) +
  labs(title = "Cell CCA — canonical correlations",
       y     = "Canonical correlation") +
  theme_classic()
```

### CSP loadings

#### Unregressed

```{r csp-loadings-cell-unreg-1}
plot_csp_loadings(cell_cca_unreg$all, cc_idx = 1)
```

```{r csp-loadings-cell-unreg-2}
plot_csp_loadings(cell_cca_unreg$all, cc_idx = 2)
```

#### Regressed

```{r csp-loadings-cell-reg-1}
plot_csp_loadings(cell_cca_reg$all, cc_idx = 1)
```

```{r csp-loadings-cell-reg-2}
plot_csp_loadings(cell_cca_reg$all, cc_idx = 2)
```

### CEP loadings

#### Unregressed

```{r cep-loadings-cell-unreg-1}
plot_cep_loadings(cell_cca_unreg$all, cc_idx = 1)
```

```{r cep-loadings-cell-unreg-2}
plot_cep_loadings(cell_cca_unreg$all, cc_idx = 2)
```

#### Regressed

```{r cep-loadings-cell-reg-1}
plot_cep_loadings(cell_cca_reg$all, cc_idx = 1)
```

```{r cep-loadings-cell-reg-2}
plot_cep_loadings(cell_cca_reg$all, cc_idx = 2)
```

### CSP-binned boundary montage

#### Unregressed

```{r montage-cell-unreg, fig.height = 3.5}
plot_csp_boundary_montage(
  group_result           = cell_cca_unreg$all,
  tpca_result            = tpca_cell_info,
  cc_idx                 = 1L,
  num_bins               = 10L,
  num_cells_for_boundary = 5L,
  norm_rescale           = TRUE
)
```

#### Regressed

```{r montage-cell-reg, fig.height = 3.5}
plot_csp_boundary_montage(
  group_result           = cell_cca_reg$all,
  tpca_result            = tpca_cell_info,
  cc_idx                 = 1L,
  num_bins               = 10L,
  num_cells_for_boundary = 5L,
  norm_rescale           = TRUE
)
```

---

# Multi-cell-type analysis

The single-cell-type sections above analyse keratinocytes in isolation. Here
we extend the workflow to multiple cell types simultaneously — endothelial
cells, fibroblasts, and keratinocytes — running NMF and TPCA per group and
then applying covariate-regressed CCA to each. This is the recommended
approach when you want to compare morpho-molecular axes across cell types
within a single experiment.

Because each cell type has its own gene programme and shape distribution,
NMF factors and shape PCs are computed **separately per group**; CCA is then
run on each group's matched shape and expression matrices. The covariate
regression helper `.regress_lmer()` defined earlier is reused without
modification.

## Step 1 — Load data and run per-group NMF (eval=FALSE)

```{r multi-load, eval = FALSE}
library(Seurat)
library(kstitch)
library(lme4)

xenium_data_path <- "/path/to/Xenium_Prime_Human_Skin_FFPE_outs/"

vignette_data <- readRDS(system.file("extdata", "vignette_part1.rds",
                                     package = "kstitch"))
cell_anno_df              <- vignette_data$Cell_Type_ID_Anno
genes_to_use_list         <- vignette_data$Genes_To_Use
cells_with_single_nucleus <- vignette_data$Cells_With_Single_Nucleus

melanoma_xenium_obj <- LoadXenium(xenium_data_path, molecule.coordinates = FALSE)
melanoma_xenium_obj <- AddMetaData(melanoma_xenium_obj, cell_anno_df)

cell_types_to_subset <- c("Endothelial", "Fibroblasts", "Keratinocytes")
subset_obj <- subset(melanoma_xenium_obj,
                     cell_type %in% cell_types_to_subset)
# You may receive object-validation warnings; these can be ignored.
```

`compute_nmf()` is called with `group.by = "cell_type"` and a named list of
per-group gene sets. It returns a named list (one entry per group) whose
`NMF_Matrix` slot contains the per-cell factor scores used by CCA.

```{r multi-nmf, eval = FALSE}
nmf_output <- compute_nmf(
  subset_obj,
  group.by                 = "cell_type",
  features                 = genes_to_use_list[cell_types_to_subset],
  assay_name               = "Xenium",
  stability_seed           = 235,
  default_num_factors      = 10,
  num_top_genes_per_factor = 50,
  integration_method       = "online",
  selectGenes_params       = list(nGenes = "all", verbose = TRUE),
  gene_pct_threshold       = 0.01
)
```

### Load segmentation info

```{r multi-seg, eval = FALSE}
melanoma_xenium_segmentation_info <- ReadXenium(
  xenium_data_path,
  outs = c("segmentation_method"),
  type = c("segmentations", "nucleus_segmentations", "centroids")
)
```

### Shared covariate vectors

```{r multi-covariates, eval = FALSE}
meta           <- subset_obj@meta.data
log_depth_vec  <- setNames(log(meta$nCount_Xenium),  rownames(meta))
seg_method_vec <- setNames(meta$segmentation_method, rownames(meta))

groups         <- cell_types_to_subset

cell_groups              <- cell_anno_df$cell_type
names(cell_groups)       <- rownames(cell_anno_df)
cell_groups_subset       <- cell_groups[cell_groups %in% cell_types_to_subset]
```

## Step 2 — Export cell contours and run TPCA (eval=FALSE)

`export_seurat_contours()` accepts a `cell_groups` named vector so it can
partition cells by type and run a separate PGA per group. The result is a
named list matching `groups`.

```{r multi-tpca-cell, eval = FALSE}
cells_to_use <- Cells(subset_obj)

tpca_cell_out <- export_seurat_contours(
  melanoma_xenium_obj,
  cell_groups  = cell_groups_subset,
  seg_list     = melanoma_xenium_segmentation_info,
  output_dir   = "~/Downloads/For_Vignette_Many_Cell_Types_TPCA",
  contour_type = "cell",
  use_parallel = TRUE,
  num_threads  = 8,
  cell_ids     = cells_to_use
)
```

## Step 3 — Cell covariate regression and CCA (eval=FALSE)

For each group we build the (shape PCs + scaled log area) matrix, regress out
sequencing depth and segmentation method from both shape and expression
matrices, then run CCA on the residuals.

```{r multi-cell-reg-cca, eval = FALSE}
cca_cell_results <- lapply(groups, function(grp) {

  # ── Cells in this group that have both TPCA and NMF results ──────────────
  tpca_cells <- rownames(tpca_cell_out[[grp]]$TPCA_Embedding)
  nmf_cells  <- rownames(nmf_output[[grp]]$NMF_Matrix)
  shared     <- intersect(tpca_cells, nmf_cells)

  # ── Scale covariates to the shared cell set ───────────────────────────────
  scaled_log_depth <- setNames(
    scale(log_depth_vec[shared])[, 1], shared
  )

  # ── Build shape matrix: top 10 shape PCs + scaled log area ───────────────
  shape_pcs    <- tpca_cell_out[[grp]]$TPCA_Embedding[shared, 1:10]
  log_area_vec <- setNames(
    log(tpca_cell_out[[grp]]$Metadata$area[
          match(shared, tpca_cell_out[[grp]]$Metadata$cell_id)]),
    shared
  )
  scaled_log_area <- setNames(scale(log_area_vec)[, 1], shared)

  shape_mat_unreg <- cbind(shape_pcs,
                           scaled_log_area = scaled_log_area)

  # ── Expression matrix ─────────────────────────────────────────────────────
  expr_mat <- nmf_output[[grp]]$NMF_Matrix[shared, ]

  # ── Regress technical covariates from both matrices ───────────────────────
  shape_mat_reg <- .regress_lmer(shape_mat_unreg, scaled_log_depth,
                                 seg_method_vec)
  expr_mat_reg  <- .regress_lmer(expr_mat,        scaled_log_depth,
                                 seg_method_vec)

  # ── CCA on regressed matrices ─────────────────────────────────────────────
  res <- link_shape_and_factors(
    obj       = subset_obj,
    shape_mat = shape_mat_reg,
    expr_mat  = expr_mat_reg,
    group.by  = NULL
  )
  res[["all"]]  # unwrap single-group result
})
names(cca_cell_results) <- groups
```

## Step 4 — Export nucleus contours and run TPCA (eval=FALSE)

Nuclear TPCA is restricted to cells with exactly one segmented nucleus.

```{r multi-tpca-nucleus, eval = FALSE}
nuclei_to_use        <- intersect(cells_to_use, cells_with_single_nucleus)
nuclei_groups_subset <- cell_groups_subset[nuclei_to_use]

tpca_nucleus_out <- export_seurat_contours(
  melanoma_xenium_obj,
  cell_groups  = nuclei_groups_subset,
  seg_list     = melanoma_xenium_segmentation_info,
  output_dir   = "~/Downloads/For_Vignette_Many_Cell_Types_TPCA",
  contour_type = "nucleus",
  use_parallel = TRUE,
  num_threads  = 8,
  cell_ids     = nuclei_to_use
)
```

## Step 5 — Nucleus covariate regression and CCA (eval=FALSE)

```{r multi-nucleus-reg-cca, eval = FALSE}
cca_nucleus_results <- lapply(groups, function(grp) {

  # ── Cells in this group that have both TPCA and NMF results ──────────────
  tpca_cells <- rownames(tpca_nucleus_out[[grp]]$TPCA_Embedding)
  nmf_cells  <- rownames(nmf_output[[grp]]$NMF_Matrix)
  shared     <- intersect(tpca_cells, nmf_cells)

  # ── Scale covariates to the shared cell set ───────────────────────────────
  scaled_log_depth <- setNames(
    scale(log_depth_vec[shared])[, 1], shared
  )

  # ── Build shape matrix: top 10 shape PCs + scaled log area ───────────────
  shape_pcs    <- tpca_nucleus_out[[grp]]$TPCA_Embedding[shared, 1:10]
  log_area_vec <- setNames(
    log(tpca_nucleus_out[[grp]]$Metadata$area[
          match(shared, tpca_nucleus_out[[grp]]$Metadata$cell_id)]),
    shared
  )
  scaled_log_area <- setNames(scale(log_area_vec)[, 1], shared)

  shape_mat_unreg <- cbind(shape_pcs,
                           scaled_log_area = scaled_log_area)

  # ── Expression matrix ─────────────────────────────────────────────────────
  expr_mat <- nmf_output[[grp]]$NMF_Matrix[shared, ]

  # ── Regress technical covariates from both matrices ───────────────────────
  shape_mat_reg <- .regress_lmer(shape_mat_unreg, scaled_log_depth,
                                 seg_method_vec)
  expr_mat_reg  <- .regress_lmer(expr_mat,        scaled_log_depth,
                                 seg_method_vec)

  # ── CCA on regressed matrices ─────────────────────────────────────────────
  res <- link_shape_and_factors(
    obj       = subset_obj,
    shape_mat = shape_mat_reg,
    expr_mat  = expr_mat_reg,
    group.by  = NULL
  )
  res[["all"]]  # unwrap single-group result
})
names(cca_nucleus_results) <- groups
```

## Multi-cell-type results

The plotting functions used in the single-cell-type analyses work identically
here: just index into `cca_cell_results` or `cca_nucleus_results` by group
name and supply the matching `tpca_result` list. The chunks below show
canonical correlation barplots across all three cell types, followed by CSP
loading and boundary montage plots for the first canonical component of each
group. Extend or subset as needed.

### Canonical correlations — cell shape

Plotting all three groups side by side makes it easy to compare the strength
of morpho-molecular coupling across cell types. Because NMF factors and shape
PCs differ between groups, the absolute correlation values are not directly
comparable, but the within-group decay across components is informative.

```{r multi-cc-cell, fig.height = 4, eval = FALSE}
cc_cell_df <- do.call(rbind, lapply(groups, function(grp) {
  coefs <- cca_cell_results[[grp]]$CC_Corr_Coefs
  data.frame(
    Group       = grp,
    Component   = paste0("CC", seq_along(coefs)),
    Correlation = coefs
  )
}))

ggplot(cc_cell_df, aes(x = Component, y = Correlation, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Cell CCA — canonical correlations by cell type",
       y     = "Canonical correlation") +
  theme_classic()
```

### Canonical correlations — nuclear shape

```{r multi-cc-nucleus, fig.height = 4, eval = FALSE}
cc_nuc_df <- do.call(rbind, lapply(groups, function(grp) {
  coefs <- cca_nucleus_results[[grp]]$CC_Corr_Coefs
  data.frame(
    Group       = grp,
    Component   = paste0("CC", seq_along(coefs)),
    Correlation = coefs
  )
}))

ggplot(cc_nuc_df, aes(x = Component, y = Correlation, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Nucleus CCA — canonical correlations by cell type",
       y     = "Canonical correlation") +
  theme_classic()
```

### Per-group CSP and CEP loadings — cell shape (CC1)

```{r multi-csp-cell, fig.height = 4, eval = FALSE}
invisible(lapply(groups, function(grp) {
  p_csp <- plot_csp_loadings(cca_cell_results[[grp]], cc_idx = 1) +
    ggtitle(sprintf("%s — cell CSP1 loadings (regressed)", grp))
  p_cep <- plot_cep_loadings(cca_cell_results[[grp]], cc_idx = 1) +
    ggtitle(sprintf("%s — cell CEP1 loadings (regressed)", grp))
  print(p_csp)
  print(p_cep)
}))
```

### Per-group CSP and CEP loadings — nuclear shape (CC1)

```{r multi-csp-nucleus, fig.height = 4, eval = FALSE}
invisible(lapply(groups, function(grp) {
  p_csp <- plot_csp_loadings(cca_nucleus_results[[grp]], cc_idx = 1) +
    ggtitle(sprintf("%s — nucleus CSP1 loadings (regressed)", grp))
  p_cep <- plot_cep_loadings(cca_nucleus_results[[grp]], cc_idx = 1) +
    ggtitle(sprintf("%s — nucleus CEP1 loadings (regressed)", grp))
  print(p_csp)
  print(p_cep)
}))
```

### Per-group boundary montages — cell shape (CC1)

```{r multi-montage-cell, fig.height = 3.5, eval = FALSE}
invisible(lapply(groups, function(grp) {
  tpca_info <- list(
    Metadata       = tpca_cell_out[[grp]]$Metadata,
    TPCA_Embedding = tpca_cell_out[[grp]]$TPCA_Embedding,
    Info           = tpca_cell_out[[grp]]$Info,
    contour_type   = "cell",
    output_dir     = NA_character_
  )
  p <- plot_csp_boundary_montage(
    group_result           = cca_cell_results[[grp]],
    tpca_result            = tpca_info,
    cc_idx                 = 1L,
    num_bins               = 10L,
    num_cells_for_boundary = 5L,
    norm_rescale           = TRUE
  )
  print(p + ggtitle(sprintf("%s — cell shape montage, CC1 (regressed)", grp)))
}))
```

### Per-group boundary montages — nuclear shape (CC1)

```{r multi-montage-nucleus, fig.height = 3.5, eval = FALSE}
invisible(lapply(groups, function(grp) {
  tpca_info <- list(
    Metadata       = tpca_nucleus_out[[grp]]$Metadata,
    TPCA_Embedding = tpca_nucleus_out[[grp]]$TPCA_Embedding,
    Info           = tpca_nucleus_out[[grp]]$Info,
    contour_type   = "nucleus",
    output_dir     = NA_character_
  )
  p <- plot_csp_boundary_montage(
    group_result           = cca_nucleus_results[[grp]],
    tpca_result            = tpca_info,
    cc_idx                 = 1L,
    num_bins               = 10L,
    num_cells_for_boundary = 5L,
    norm_rescale           = TRUE
  )
  print(p + ggtitle(sprintf("%s — nucleus shape montage, CC1 (regressed)", grp)))
}))
```

See `?export_seurat_contours`, `?compute_nmf`, and `?link_shape_and_factors`
for full argument documentation.
